---
title: "Quesnel NRD connectivity"
subtitle: "Corrections and rerun, August 2026"
author: "Alex Chubaty"
date: last-modified
date-format: "[Last updated:] YYYY-MM-DD"
format:
  pdf:
    # Distinct from the Quesnel branch's `corrections-2026-08.pdf`. `Outputs/` is shared storage,
    # not per-branch, so without this the two versions of this report overwrite each other
    # depending on which was rendered last. Drop the suffix when the branches merge and there is
    # only one.
    output-file: corrections-2026-08-cariboo.pdf
    documentclass: article
    geometry: margin=2.5cm
    toc: true
    toc-depth: 2
    number-sections: true
    fig-pos: "H"
    pdf-engine: pdflatex
    urlcolor: blue
    include-in-header:
      text: |
        \usepackage{float}
        \floatplacement{figure}{H}
        %% Tables are floats too, and Quarto emits their `table` environment itself -- so
        %% kableExtra's `HOLD_position` has nothing left to pin. Pin them here instead, or a
        %% table drifts pages away from the paragraph that discusses it.
        \floatplacement{table}{H}
        %% kableExtra::linebreak() emits \makecell to wrap a column header onto two rows
        \usepackage{makecell}
        %% Keep a heading with the beginning of the text under it. LaTeX already refuses to break
        %% immediately after a heading, but nothing stops it breaking after the first line or two,
        %% which strands the heading at the foot of a page and reads as though the section were
        %% empty. \needspace reserves room for the heading plus a few lines and breaks the page
        %% first if it is not there.
        \usepackage{needspace}
        \let\oldsection\section
        \renewcommand{\section}{\needspace{5\baselineskip}\oldsection}
        \let\oldsubsection\subsection
        \renewcommand{\subsection}{\needspace{5\baselineskip}\oldsubsection}
        \let\oldsubsubsection\subsubsection
        \renewcommand{\subsubsection}{\needspace{5\baselineskip}\oldsubsubsection}
        %% and no single line of a paragraph left alone at the foot or head of a page
        \clubpenalty=10000
        \widowpenalty=10000
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  ## Keep figures in the section that discusses them. Captioned figures are LaTeX floats,
  ## so by default two full-width 4800 px maps drift to wherever they fit -- which put them
  ## pages away from the text about them. `H` (from the float package, loaded in the YAML
  ## header) pins them exactly where they appear in the source.
  fig.pos = "H",
  out.extra = ""
)

## knitr evaluates chunks in the directory of the input file, so anything that resolves the project
## by falling back to the working directory (`workflowtools::findProjectPath()` among them) reports
## `reports/` and every derived path silently misses. Walk up to the directory holding `_targets.R`.
find_project_root <- function(start = getwd()) {
  p <- normalizePath(start, mustWork = FALSE)
  while (!file.exists(file.path(p, "_targets.R")) && dirname(p) != p) {
    p <- dirname(p)
  }
  if (!file.exists(file.path(p, "_targets.R"))) {
    stop("could not locate the project root")
  }
  p
}

proj <- find_project_root()
knitr::opts_knit$set(root.dir = proj)

## Emit native LaTeX tables rather than markdown ones. Pandoc turns a markdown table into a
## `longtable`, which splits across pages -- so a table lands half on one page and half on the next.
## `format = "latex"` with `booktabs` produces a `table` float instead, and floats never split; the
## preamble then pins them in place and sets them a size down.
options(knitr.table.format = "latex")

library(dplyr)
library(knitr)
library(targets)

## The pipeline is run once per Natural Resource District, and each district's inputs and results
## live under its own key. Quesnel is one district among them; nothing here is flat.
QUESNEL_OUT <- file.path(proj, "Outputs", "quesnel")

## Artefacts of the delivered (pre-correction) run, copied out of the way before the rerun
## overwrote them. See Appendix C, "Reproducing this report".
##
## The directory keeps the name it was given at the time; only the wording in this report changed.
DELIVERED <- file.path(proj, "Data", "processed", "quesnel", "published-2026-01-27")
## Derived from wherever `_targets` actually resolves to (it is usually a symlink to bulk storage),
## so no site-specific absolute path is baked into the repo. Override with the environment variable.
DELIVERED_META <- Sys.getenv(
  "BC_CONN_DELIVERED_META",
  file.path(
    dirname(normalizePath(file.path(proj, "_targets"), mustWork = FALSE)),
    "_targets_meta_backup_pre-2026-08-25-fixes"
  )
)
## Figures live beside this document, wherever it is. Building the path out of `proj` instead
## hardcodes the assumption that this file sits in `reports/`, and when project-root resolution
## slipped that assumption produced a nested `reports/reports/figures`.
rmd_dir <- tryCatch(dirname(knitr::current_input(dir = TRUE)), error = function(e) NULL)

if (is.null(rmd_dir) || !nzchar(rmd_dir) || is.na(rmd_dir)) {
  rmd_dir <- file.path(proj, "reports") ## sourcing the chunks outside knitr
}

FIGDIR <- normalizePath(file.path(rmd_dir, "figures"), mustWork = FALSE)
dir.create(FIGDIR, showWarnings = FALSE, recursive = TRUE)

## Pandoc resolves an image path against the directory of the input file, and an ABSOLUTE path
## reaches LaTeX with a `.` glued to the front: `./home/achubaty/.../patches_delivered.png`, which
## does not exist and fails the build. rmarkdown did not do this, so the .Rmd could hand it the
## absolute path. Keep the absolute path for the existence check in R, which runs with the project
## root as its working directory, and give pandoc the relative one.
fig_rel <- function(f) file.path(basename(FIGDIR), basename(f))

## One place that decides how every table in this report is drawn.
##
## It draws the table body only. The caption, the number and the label come from the chunk's
## `#| tbl-cap` and `#| label: tbl-...` options, so that every table can be referred to as
## `@tbl-...` and no number in the prose can go stale. This also removes a whole class of defect:
## a caption written here was passed through to LaTeX unescaped, and a bare `%` in one silently
## centred every paragraph from that table to the end of the document. Captions handled by pandoc
## are escaped properly.
##
## `striped` shades alternate rows, which is what makes a table of more than a few rows readable
## across its full width.
##
## `widths` is a per-column width, e.g. c("8cm", "7cm"). A column given one wraps inside it;
## without one, a long cell sets the column to its own length and pushes the table off the page.
##
## Cell contents are plain text. `kbl()` escapes them, so they must not carry markdown: a backtick
## or a pair of asterisks would reach the page literally. Where a cell needs monospace or bold,
## build it with `kableExtra::cell_spec()` rather than marking it up by hand.
## `stripe_by` shades alternate *groups* of rows rather than alternate rows, for a table whose
## rows come in blocks: banding every second row across a block obscures the blocks, which are
## the thing the reader is trying to follow.
styled <- function(x, ..., widths = NULL, bold_rows = NULL, stripe_by = NULL) {
  latex_opts <- if (is.null(stripe_by)) "striped" else NULL

  ## `booktabs` inserts a gap after every fifth row by default, which lands wherever it lands: in
  ## a twelve-row table of four three-row blocks it fell inside the second and fourth blocks and
  ## read as though it meant something. Suppress it, and where the rows are grouped put the gap on
  ## the group boundaries instead, where it reinforces the shading rather than cutting across it.
  linesep <- ""
  if (!is.null(stripe_by)) {
    g <- match(stripe_by, unique(stripe_by))
    linesep <- ifelse(c(diff(g), 1) != 0, "\\addlinespace", "")
    linesep[length(linesep)] <- "" ## no gap between the last row and the bottom rule
  }

  k <- kableExtra::kbl(x, linesep = linesep, ...) |>
    kableExtra::kable_styling(latex_options = latex_opts, font_size = 9)

  if (!is.null(stripe_by)) {
    g <- match(stripe_by, unique(stripe_by))
    shaded <- which(g %% 2L == 0L)
    if (length(shaded)) {
      k <- kableExtra::row_spec(k, shaded, background = "#EFEFEF")
    }
  }

  for (i in seq_along(widths)) {
    if (nzchar(widths[i])) {
      k <- kableExtra::column_spec(k, i, width = widths[i])
    }
  }
  if (!is.null(bold_rows)) {
    k <- kableExtra::row_spec(k, bold_rows, bold = TRUE)
  }
  k
}

## Emit a table into a `results: asis` chunk.
##
## The surrounding blank lines are load-bearing: without them pandoc can read the table's LaTeX as
## a continuation of the preceding paragraph and start escaping parts of it into visible text while
## leaving others live.
emit <- function(k) cat("\n\n", as.character(k), "\n\n", sep = "")

ha <- function(x) sum(terra::expanse(x, unit = "ha", transform = FALSE))
num <- function(x, d = 1) formatC(round(x, d), format = "f", big.mark = ",", digits = d)

## Patch areas span five orders of magnitude within a single column, so a fixed number of decimal
## places is either unreadable at the top or meaningless at the bottom. Scientific notation keeps
## every row the same width and the same precision.
sci <- function(x, d = 2) formatC(x, format = "e", digits = d)
dur <- function(s) {
  ifelse(
    is.na(s),
    "n/a",
    ifelse(
      s >= 3600,
      sprintf("%.2f h", s / 3600),
      ifelse(s >= 60, sprintf("%.1f min", s / 60), sprintf("%.1f s", s))
    )
  )
}
have <- function(...) all(file.exists(...))

## `tar_read()` happily returns a target that has NOT been rebuilt: the store keeps the delivered
## run's object and metadata until the target is remade. Reading one of those and calling it
## "corrected" would silently compare the delivered run against itself. A target counts as rebuilt
## only where its recorded time differs from the delivered snapshot.
secs_by_name <- function(store) {
  if (!dir.exists(store)) {
    return(NULL)
  }
  m <- tar_meta(store = store, fields = c("name", "seconds"))
  stats::setNames(m$seconds, m$name)
}

PUB_SECONDS <- secs_by_name(DELIVERED_META)
CUR_SECONDS <- secs_by_name(file.path(proj, "_targets"))

is_rebuilt <- function(target) {
  if (is.null(PUB_SECONDS) || !target %in% names(PUB_SECONDS)) {
    return(TRUE)
  }
  if (is.null(CUR_SECONDS) || !target %in% names(CUR_SECONDS)) {
    return(FALSE)
  }
  cur <- CUR_SECONDS[[target]]
  if (is.na(cur)) {
    return(FALSE)
  }
  !isTRUE(all.equal(cur, unname(PUB_SECONDS[[target]])))
}
```

{{< pagebreak >}}

# Summary {#sec-summary}

An issue found in the seral-patch construction (commit `145cffb`) turned out to be one of several, and together they invalidate the connectivity results delivered in January 2026.
This report documents what was wrong, what the corrected numbers are, what the rerun cost, and why it needed to be rerun.

Omniscape reads two layers: the composite resistance surface and the composite source weights.
Both change materially.
Roughly half of what the delivered run treated as old interior forest was not forest at all, and a further large block of genuine old forest was modelled at the wrong resistance.
The Omniscape runs therefore had to be repeated rather than patched.

A second strand of work addressed the two largest spatial overlays.
Tiling the final resultant is done and took it from 5 h 52 min to 18 min.
Tiling the interior-forest overlay took it from 3 h 08 min to 6 min, reproducing the stored output exactly.
That matters here because the corrections below had to be applied, checked and re-applied several
times over, and a step measured in hours is one that cannot be retried casually.
It matters more for what comes next: the workflow is intended to cover the wider Cariboo, and
Cariboo-Chilcotin alone is roughly 2.4 times Quesnel's area, with these costs growing faster than
area rather than in step with it.
The retooling is a prerequisite for that extension rather than a refinement of it.

```{r}
#| label: tbl-summary-effects
#| tbl-cap: "Principal effects of the corrections."
data.frame(
  Item = c(
    "Seral layer, total area",
    "Old interior forest",
    "VRI x NDT-BEC layer, double-counted area",
    "Pipeline steps eliminated"
  ),
  Delivered = c("2,931,558 ha", "326,081.5 ha (13,783 polys)", "15,449.2 ha", "n/a"),
  Corrected = c("see Forest patches, below", "166,915.6 ha (10,348 polys)", "0.0 ha", "35.1 h of compute"),
  stringsAsFactors = FALSE
) |>
  styled(booktabs = TRUE)
```

# Issues found and corrected {#sec-issues}

None of these produced an error.
Each produced numbers that looked reasonable, which is why they survived into the delivered results.

```{r}
#| label: tbl-issues
#| tbl-cap: "Issues corrected in this rerun, in the order the pipeline applies them. The numbering is referred to by the subsections that follow."
data.frame(
  `#` = seq_len(8),
  Issue = c(
    "The buffer that defines interior forest was removed from the mature-and-old class as well as from the old class",
    "Patches were not restricted to those actually classified as interior forest",
    "Interior forest overwrote the seral stage it was derived from",
    "Seral stages were assigned by spatial overlap rather than by splitting polygons at their boundaries",
    "The leading-species layer was combined with a tool that allows its outputs to overlap",
    "Slivers were removed by area rather than by the shared-edge rule the original scripts use",
    "Patch statistics were measured on overlay fragments rather than on whole patches",
    "Tile overlap for Omniscape was given in pixels where map units were expected"
  ),
  Consequence = c(
    "Mature-and-old interior forest came out identical to old-only interior forest",
    "50.8% of the area reported as old interior forest was unclassified land",
    "160,417 ha of old forest cores were modelled as ordinary forest instead of as the best available habitat",
    "The layer grew to 7.78 million polygons covering 39.5 million ha, far more than the study area holds",
    "22,628 ha of land was dropped and 15,449 ha was counted twice",
    "Slivers were merged into the wrong neighbouring polygon",
    "The smallest reported old patch was < 4.6e-6 m\u00b2",
    "Visible seams across the tiled test run"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
) |>
  styled(
    booktabs = TRUE,
    row.names = FALSE,
    widths = c("0.6cm", "7.2cm", "7.2cm")
  )
```

The rest of this section gives the detail behind each, and the area figures that follow from them.
Issue 8 concerns how the Omniscape runs were tiled for benchmarking rather than the data itself, and is covered in [Appendix B](#sec-appendix-timings).

## Forest patches {#sec-forest-patches}

*Issues 1, 2, 3, 4 and 6.*

```{r patches-table}
pub_gpkg <- file.path(DELIVERED, "forest_patches_final.gpkg")
ok_patches <- have(pub_gpkg) &&
  !inherits(try(tar_read_raw("forest_patches_final"), silent = TRUE), "try-error") &&
  is_rebuilt("forest_patches_final")
```

```{r}
#| label: tbl-patches-compare
#| tbl-cap: "Final resultant by seral stage. Resistance and source weight are assigned from the seral stage, so a stage that changes area moves that area between resistance classes."
#| eval: !expr ok_patches
tabulate_patches <- function(v) {
  v$ha <- terra::expanse(v, unit = "ha", transform = FALSE)
  as.data.frame(v) |>
    group_by(Seral, Resistance, SourceWt) |>
    summarise(n = n(), ha = sum(ha), .groups = "drop")
}

b <- terra::vect(pub_gpkg)
a <- tidyterra::as_spatvector(tar_read_raw("forest_patches_final"))

cmp <- full_join(
  tabulate_patches(b) |> rename(n_pub = n, ha_pub = ha),
  tabulate_patches(a) |> rename(n_cor = n, ha_cor = ha),
  by = c("Seral", "Resistance", "SourceWt")
) |>
  mutate(
    across(where(is.numeric), \(x) ifelse(is.na(x), 0, x)),
    d = ha_cor - ha_pub,
    pct = ifelse(ha_pub > 0, 100 * d / ha_pub, NA_real_)
  ) |>
  arrange(Resistance)

cmp |>
  transmute(
    Seral,
    Resistance,
    SourceWt,
    `Delivered (ha)` = num(ha_pub),
    `Corrected (ha)` = num(ha_cor),
    `Change (ha)` = num(d),
    `Change (%)` = num(pct)
  ) |>
  styled(
    booktabs = TRUE
  )
```

```{r patches-missing, eval = !ok_patches, results = "asis"}
cat("_The corrected `forest_patches_final` was not available when this was rendered._\n")
```

@fig-patches-delivered, @fig-patches-corrected and @fig-patches-change are drawn by the same code
from the same inputs, so any difference between them is a difference in the data rather than in how
it was plotted.

```{r}
#| label: fig-patches-delivered
#| fig-cap: "Delivered patches, January 2026. Old forest is thin and fragmented; mature forest is near-continuous."
#| eval: !expr ok_patches
#| out-width: "100%"
targets::tar_source()
f <- file.path(FIGDIR, "patches_delivered.png")
## the writers return the path they wrote; at the top level of a chunk an `if` whose condition
## holds prints its value, which puts the filename in the rendered document
if (!file.exists(f)) {
  invisible(plot_forest_disturbance_seral(
    b,
    tar_read_raw("Quesnel_TSA"),
    "patches_delivered.png",
    outdir = FIGDIR
  ))
}
include_graphics(fig_rel(f), rel_path = FALSE, error = FALSE)
```

```{r}
#| label: fig-patches-corrected
#| fig-cap: "Corrected patches, August 2026."
#| eval: !expr ok_patches
#| out-width: "100%"
f <- file.path(FIGDIR, "patches_corrected.png")
if (!file.exists(f)) {
  invisible(plot_forest_disturbance_seral(
    a,
    tar_read_raw("Quesnel_TSA"),
    "patches_corrected.png",
    outdir = FIGDIR
  ))
}
include_graphics(fig_rel(f), rel_path = FALSE, error = FALSE)
```

Set side by side, the two maps show that the corrected old forest is less extensive, but not where
the area went.
@fig-patches-change answers that: for each seral stage it separates the ground both runs agree on
from the ground only one of them claims.

```{r}
#| label: fig-patches-change
#| fig-cap: "Change in each seral stage. Grey is ground both runs assign to the stage; orange is ground only the delivered run assigned to it, and blue only the corrected run."
#| eval: !expr ok_patches
#| out-width: "100%"
## Both vintages have to land on ONE grid. `seral_display_raster()` builds its template from the
## extent of whatever it is handed, so rasterising the two separately would put them on grids that
## do not line up, and a cell-by-cell comparison of those means nothing.
seral_change_map <- function(pub, cor, studyArea, dst, res = 100) {
  vp <- tidyterra::as_spatvector(pub)
  vc <- tidyterra::as_spatvector(cor)

  ep <- terra::ext(vp)
  ec <- terra::ext(vc)
  e <- terra::ext(
    min(ep[1], ec[1]), max(ep[2], ec[2]),
    min(ep[3], ec[3]), max(ep[4], ec[4])
  )
  template <- terra::rast(e, resolution = res, crs = terra::crs(vp))

  ## Unclassified stands are carried as an explicit level for the same reason the single-vintage
  ## map carries them: a rasterised NA cannot be told apart from a cell no polygon covers.
  burn <- function(v) {
    v$.display <- dplyr::coalesce(as.character(v$Seral), SERAL_UNCLASSIFIED)
    terra::rasterize(v, template, field = ".display")
  }

  d <- terra::as.data.frame(burn(vp), xy = TRUE, na.rm = FALSE)
  names(d)[3] <- "pub"
  d$cor <- terra::as.data.frame(burn(vc), xy = TRUE, na.rm = FALSE)[[3]]

  ## stand-development order, not alphabetical: Mature sorts before Mid and reads as an error
  classes <- intersect(SERAL_LEVELS, unique(c(d$pub, d$cor)))
  long <- do.call(rbind, lapply(classes, function(k) {
    in_pub <- !is.na(d$pub) & d$pub == k
    in_cor <- !is.na(d$cor) & d$cor == k
    keep <- in_pub | in_cor
    data.frame(
      x = d$x[keep],
      y = d$y[keep],
      Seral = factor(k, levels = classes),
      Change = ifelse(
        in_pub[keep] & in_cor[keep], "Unchanged",
        ifelse(in_cor[keep], "Added by the correction", "Removed by the correction")
      )
    )
  }))
  long$Change <- factor(
    long$Change,
    levels = c("Unchanged", "Removed by the correction", "Added by the correction")
  )

  sa <- tidyterra::as_spatvector(studyArea) |>
    terra::project(terra::crs(template)) |>
    sf::st_as_sf()

  g <- ggplot2::ggplot(long) +
    ggplot2::geom_tile(ggplot2::aes(x = .data$x, y = .data$y, fill = .data$Change)) +
    ggplot2::facet_wrap(ggplot2::vars(.data$Seral), ncol = 2) +
    ggplot2::geom_sf(data = sa, color = "black", fill = NA, inherit.aes = FALSE) +
    ggplot2::scale_fill_manual(
      values = c(
        "Unchanged" = "grey80",
        "Removed by the correction" = "#d95f0e",
        "Added by the correction" = "#2c7fb8"
      )
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(legend.position = "bottom") +
    ggspatial::annotation_north_arrow(
      location = "bl",
      which_north = "true",
      pad_x = ggplot2::unit(0.25, "in"),
      pad_y = ggplot2::unit(0.25, "in"),
      style = ggspatial::north_arrow_fancy_orienteering
    ) +
    ggplot2::xlab("Longitude") +
    ggplot2::ylab("Latitude") +
    ggplot2::ggtitle("Quesnel NRD: change in seral stage extent")

  ggplot2::ggsave(dst, g, width = 16, height = 16)
  dst
}

f <- file.path(FIGDIR, "patches_change.png")
if (!file.exists(f)) {
  invisible(seral_change_map(b, a, tar_read_raw("Quesnel_TSA"), f))
}
include_graphics(fig_rel(f), rel_path = FALSE, error = FALSE)
```

## Patch statistics {#sec-patch-statistics}

*Issue 7.*

Patch statistics now apply a 1 ha floor, matching the CEF Biodiversity Protocol threshold for
residual patches and the threshold arcpy `Eliminate` applies.
This is a reporting filter only.
It never changes the layers the connectivity analysis is built from, and the corrected run records
how many patches it excludes and how much area they hold.

The floor is visible in the first column below.
The delivered statistics were measured on overlay fragments rather than on whole patches, so the
smallest patch in every stage came out at a ten-billionth of a hectare or less, which is a sliver a
fraction of a square millimetre across.
The corrected smallest patch is exactly 1 ha in every stage, which is the floor doing its work.

```{r}
#| label: tbl-patch-stats
#| tbl-cap: "Seral patch statistics, delivered against corrected, paired by stage. All areas are hectares, given to three significant figures. Only the statistics both runs recorded are shown."
#| results: asis
pub_stats <- file.path(DELIVERED_META, "Quesnel_TSA_seral_patch_stats.csv")
cor_stats <- file.path(QUESNEL_OUT, "Quesnel_TSA_seral_patch_stats.csv")
if (have(pub_stats, cor_stats)) {
  p <- read.csv(pub_stats)
  q <- read.csv(cor_stats)

  ## Only the columns both runs recorded can be compared: the corrected run also reports patch
  ## counts, totals and exclusions, which the delivered one did not, so those are named in the
  ## prose rather than shown here as a column of blanks.
  keep <- intersect(names(p), names(q))
  labels <- c(
    Area_Min = "Smallest (ha)",
    Area_Mean = "Mean (ha)",
    Area_Median = "Median (ha)",
    Area_Max = "Largest (ha)",
    Area_Total = "Total (ha)"
  )
  area_cols <- intersect(names(labels), keep)

  ## Paired by seral stage rather than stacked by run, so the comparison the table exists to make
  ## is between adjacent rows.
  stats <- bind_rows(
    mutate(p[, keep, drop = FALSE], Run = "Delivered"),
    mutate(q[, keep, drop = FALSE], Run = "Corrected")
  ) |>
    mutate(Seral = factor(Seral, levels = SERAL_LEVELS)) |>
    arrange(Seral, match(Run, c("Delivered", "Corrected")))

  out <- stats[, c("Seral", "Run")]
  for (cl in area_cols) {
    out[[labels[[cl]]]] <- sci(stats[[cl]] / 1e4)
  }

  out |>
    styled(
      booktabs = TRUE,
      row.names = FALSE,
      align = c("l", "l", rep("r", length(area_cols))),
      stripe_by = stats$Seral
    ) |>
    emit()
} else {
  cat("_Patch statistics not available._")
}
```

## A note on the overlay tools {#sec-overlay-tools-note}

Several of the issues above came from the tools used to combine map layers rather than from the data
itself, and two of them turned out to be faults in a widely used spatial library rather than in this
pipeline.
That matters for anyone reusing the approach, but it does not change any area reported above, so the
account is in [Appendix A](#sec-appendix-overlay-tools).

# Effect on the connectivity analysis {#sec-effect}

Omniscape consumes exactly two rasters: the composite resistance surface and the composite source
weights.
Both are assigned from `Seral`, so any area that changes seral stage moves between resistance
classes, and the connectivity result changes with it.

This section compares what Omniscape read and what it produced, before and after the corrections.
Both sides are still on disk: the delivered inputs were copied aside before the rerun overwrote
them, and the delivered runs survived because the corrections changed the run names ([Appendix B](#sec-run-naming)).

## What changed in the inputs {#sec-inputs-changed}

```{r omniscape-rasters}
RAST_PUB <- file.path(DELIVERED, "rasters")
## Note the asymmetry: the per-district raster directory is `rasters/quesnel`, not
## `quesnel/rasters` -- the layout keys rasters off the shared `rasters` parent, unlike the rest.
RAST_CUR <- file.path(proj, "Data", "processed", "rasters", "quesnel")

## The corrected composite is written by the `resistance_composite` target; until that has been
## rebuilt the file on disk is still the delivered one, and comparing it with its own backup would
## report no change at all.
ok_rast <- dir.exists(RAST_PUB) && is_rebuilt("resistance_composite")

## Read each raster once into a plain vector and do the arithmetic in R. Repeated
## `terra::global()` calls re-read the file for every statistic, which at 30 m -- 69.2 M cells --
## turned this comparison from seconds into minutes.
##
## A cell counts as changed if its value differs OR its NA status differs. The second case is not
## a detail here: the forest layers are NA off-patch, so a cell that leaves the patch set goes
## from a number to NA, and a value-only comparison would miss it entirely.
cmp_rast <- function(name, res) {
  f <- sprintf("%s_%d.tif", name, res)
  ra <- terra::rast(file.path(RAST_PUB, f))
  rb <- terra::rast(file.path(RAST_CUR, f))
  if (!terra::compareGeom(ra, rb, stopOnError = FALSE)) {
    return(NULL)
  }
  va <- terra::values(ra, mat = FALSE)
  vb <- terra::values(rb, mat = FALSE)
  na_a <- is.na(va)
  na_b <- is.na(vb)
  both <- !na_a & !na_b
  cellha <- prod(terra::res(ra)) / 1e4
  out <- data.frame(
    layer = name,
    res = res,
    ha_chg = (sum(na_a != na_b) + sum(both & va != vb)) * cellha,
    ha_lost = sum(!na_a & na_b) * cellha,
    ha_gained = sum(na_a & !na_b) * cellha,
    mean_pub = mean(va[both]),
    mean_cor = mean(vb[both]),
    stringsAsFactors = FALSE
  )
  rm(va, vb, na_a, na_b, both)
  gc(verbose = FALSE)
  out
}

LAYERS <- c("resistance_forest", "resistance_composite", "sourcewt_forest", "sourcewt_composite")
LAYER_LABELS <- c(
  resistance_forest = "Resistance, forest",
  resistance_composite = "Resistance, composite",
  sourcewt_forest = "Source weight, forest",
  sourcewt_composite = "Source weight, composite"
)

## Prose below quotes these inline. Without NA defaults a missing artefact would turn a chunk that
## is merely skipped into a render that fails, several pages from the chunk that did not run.
tot_ha <- NA_real_
ca <- cb <- stats::setNames(rep(NA_real_, 6), c("1", "250", "500", "750", "1000", "mixed"))
pair_pub <- pair_cor <- NA_real_
```

```{r}
#| label: tbl-omniscape-input-diff
#| tbl-cap: "Change in the rasters Omniscape reads. The forest layers are NA off-patch; the composites fill those cells with the matrix value and are complete over the study area. The mean is of the layer's own values, taken over cells present in both versions: a resistance score from 1 to 1000, or a source weight from 0 to 1, according to the layer. It is not an area."
#| eval: !expr ok_rast
#| results: asis
inp <- do.call(
  rbind,
  lapply(c(30, 90), function(r) {
    do.call(rbind, lapply(LAYERS, cmp_rast, res = r))
  })
)

## Study-area total, for the share-changed column. Taken from the composite, which unlike the
## forest layers is complete over the study area -- it has no NA cells at all.
tot_ha <- {
  r <- terra::rast(file.path(RAST_CUR, "resistance_composite_30.tif"))
  sum(!is.na(terra::values(r, mat = FALSE))) * prod(terra::res(r)) / 1e4
}

## Resistance is a score from 1 to 1000 and source weight a fraction from 0 to 1, so the two kinds
## of layer want different precision in the same column.
dp <- ifelse(grepl("sourcewt", inp$layer), 4, 2)

tab <- data.frame(
  Layer = unname(LAYER_LABELS[inp$layer]),
  res = inp$res,
  ha = num(inp$ha_chg, 0),
  share = num(100 * inp$ha_chg / tot_ha, 2),
  mean_pub = mapply(num, inp$mean_pub, dp),
  mean_cor = mapply(num, inp$mean_cor, dp),
  check.names = FALSE
)

## Two-row headers keep the table inside the text width. `linebreak()` emits `\makecell`, which
## needs `escape = FALSE`, so from here the cells have to be LaTeX-safe themselves: the per-cent
## sign lives in the header, written as `\%`, and the cells carry bare numbers.
names(tab) <- kableExtra::linebreak(
  c(
    "Layer",
    "Resolution\n(m)",
    "Area changed\n(ha)",
    "Share of\nstudy area (\\%)",
    "Mean value\ndelivered",
    "Mean value\ncorrected"
  ),
  align = "c"
)

tab |>
  styled(
    booktabs = TRUE,
    row.names = FALSE,
    escape = FALSE
  ) |>
  emit()
```

```{r omniscape-pending, eval = !ok_rast, results = "asis"}
cat("_The corrected resistance surface had not been rebuilt when this was rendered._\n")
```

The 30 m and 90 m rows are independent measurements of the same change on the ground, and they
agree to better than 0.1% on every layer.
That is the check that this is a real difference in the patch set and not an artefact of one grid.

Two things in @tbl-omniscape-input-diff need separating.

**The composite means move in the direction of less connectivity**: mean resistance rises about
1.4% and mean source weight falls about 4.8%.
The corrected landscape is slightly harder to move through than the delivered one said it was.

**The forest layers change over more ground than the composites do**, which is not a contradiction.
The forest layer is NA wherever there is no patch, so it registers a change both when a cell's
resistance class moves and when the cell enters or leaves the patch set; in the composite the
off-patch cells carry the matrix value instead, so only the first kind appears there as a change.

```{r}
#| label: tbl-omniscape-extent
#| tbl-cap: "Change in the extent of the forest patch layer, at 30 m."
#| eval: !expr ok_rast
#| results: asis
fe <- subset(inp, layer == "resistance_forest" & res == 30)
data.frame(
  Measure = c(
    "In the delivered patch layer, not the corrected one",
    "In the corrected patch layer, not the delivered one",
    "Net change in patch extent"
  ),
  Value = c(
    paste(num(fe$ha_lost, 0), "ha"),
    paste(num(fe$ha_gained, 0), "ha"),
    paste(num(fe$ha_gained - fe$ha_lost, 0), "ha")
  ),
  stringsAsFactors = FALSE
) |>
  styled(
    booktabs = TRUE
  ) |>
  emit()
```

The loss is almost entirely one-directional, and it is the raster expression of the patch-area
change reported under "Forest patches".

### Where the land went {#sec-where-the-land-went}

Resistance is assigned in discrete classes, so the composite can be compared class by class.

```{r}
#| label: tbl-omniscape-classes
#| tbl-cap: "Area in each resistance class of the composite at 30 m, with the source weight assigned from the same seral class. Mixed cells are those covered by more than one polygon, which rasterize averages."
#| eval: !expr ok_rast
#| results: asis
cellha <- 30 * 30 / 1e4

## The nominal classes are what `Resistance` is assigned from. Everything else in the raster is a
## cell that `terra::rasterize(fun = mean)` averaged because more than one polygon covered it,
## and those are pooled into one row -- they are the subject of the paragraph below.
##
## `terra::freq()` counts values in C++ without materialising 69.2 M doubles in R, which matters
## here because the transition table below needs both vectors in memory at once anyway.
##
## `digits` defaults to 0, which ROUNDS the values before counting -- it silently merges 125.5 into
## 126, and the mixture values this section is about stop existing. The nominal classes are whole
## numbers so the totals come out right either way, which is what makes the default dangerous here.
NOMINAL <- c(1, 250, 500, 750, 1000)
FREQ_DIGITS <- 4
cls <- function(f) {
  tb <- terra::freq(terra::rast(f), digits = FREQ_DIGITS)
  k <- ifelse(tb$value %in% NOMINAL, as.character(tb$value), "mixed")
  out <- tapply(tb$count * cellha, factor(k, levels = c(as.character(NOMINAL), "mixed")), sum)
  out[is.na(out)] <- 0
  out
}
ca <- cls(file.path(RAST_PUB, "resistance_composite_30.tif"))
cb <- cls(file.path(RAST_CUR, "resistance_composite_30.tif"))

## The two mixture values that are exactly the average of an adjacent pair of seral classes:
## 125.5 = (1 + 250) / 2 and 375 = (250 + 500) / 2. They are the ones the paragraph below is about.
pair_ha <- function(f) {
  tb <- terra::freq(terra::rast(f), digits = FREQ_DIGITS)
  sum(tb$count[abs(tb$value - 125.5) < 1e-6 | abs(tb$value - 375) < 1e-6]) * cellha
}
pair_pub <- pair_ha(file.path(RAST_PUB, "resistance_composite_30.tif"))
pair_cor <- pair_ha(file.path(RAST_CUR, "resistance_composite_30.tif"))

data.frame(
  `Resistance` = c(paste0(NOMINAL, c(" (best)", "", "", "", " (matrix)")), "mixed cells"),
  `Source weight` = c("1.00", "0.75", "0.50", "0.25", "0.00", "n/a"),
  `Delivered (ha)` = num(as.numeric(ca), 0),
  `Corrected (ha)` = num(as.numeric(cb), 0),
  `Change (ha)` = paste0(ifelse(cb - ca >= 0, "+", ""), num(as.numeric(cb - ca), 0)),
  check.names = FALSE
) |>
  styled(
    booktabs = TRUE,
    row.names = FALSE
  ) |>
  emit()
```

The correction **polarises** the landscape.
The best class more than doubles and the matrix grows, both at the expense of the middle: the
250 class alone loses about a quarter of a million hectares, split between the two extremes.
This is why the composite means barely move while so much ground changes class: the gains at each end largely cancel.

It also explains the last row, which is worth dwelling on because it is independent evidence.

`create_resistance_raster()` rasterises with `fun = mean`, and `terra::rasterize()` assigns a cell
from every polygon covering its centre.
A cell written as 125.5 is therefore one whose centre fell inside a resistance-1 polygon **and** a
resistance-250 polygon at the same time, 125.5 being the average of the two.
Cells like that can only exist where the patch layer overlaps itself, which is precisely what the
`terra::union()` issue produced.
The composite is assembled with `pmax()` and does no averaging of its own, so every mixed cell in
it was already mixed in one of the layers it reads.

Mixed cells fall from `r num(ca[["mixed"]], 0)` ha to `r num(cb[["mixed"]], 0)` ha.
The two values that are exactly the average of an adjacent pair of seral classes, 125.5 and 375,
account for `r num(pair_pub, 0)` ha of the delivered total and `r num(pair_cor, 0)` ha of the
corrected one.
That is raster-level confirmation, arrived at without going near the polygons, that the overlapping
geometry is gone.
The `r num(cb[["mixed"]], 0)` ha that remain are `r num(100 * cb[["mixed"]] / tot_ha, 2)`% of the
study area; overlap has not been eliminated outright, but what is left is small enough not to move
any figure in this report.

```{r}
#| label: tbl-omniscape-transitions
#| tbl-cap: "The six largest resistance-class transitions in the composite at 30 m."
#| eval: !expr ok_rast
#| results: asis
## Tabulate only the cells that actually changed. A full cross-tabulation of both vectors would
## factor 69.2 M values against ~300 levels twice; restricted to the changed cells it is 18x
## smaller and the result is identical, because the diagonal is dropped anyway.
va <- terra::values(terra::rast(file.path(RAST_PUB, "resistance_composite_30.tif")), mat = FALSE)
vb <- terra::values(terra::rast(file.path(RAST_CUR, "resistance_composite_30.tif")), mat = FALSE)
ch <- which(va != vb)
tr <- as.data.frame(table(from = va[ch], to = vb[ch]), stringsAsFactors = FALSE)
tr <- tr[tr$Freq > 0, ]
tr$ha <- tr$Freq * cellha
tr <- head(tr[order(-tr$ha), ], 6)

## These are class labels, so they get no thousands separator and no forced decimal: "1000",
## not "1,000.0", and "125.5" kept as it is.
lbl <- function(x) ifelse(x == round(x), sprintf("%.0f", x), sprintf("%.1f", x))

data.frame(
  `Delivered resistance` = lbl(as.numeric(tr$from)),
  `Corrected resistance` = lbl(as.numeric(tr$to)),
  `Area (ha)` = num(tr$ha, 0),
  check.names = FALSE
) |>
  styled(
    booktabs = TRUE,
    row.names = FALSE
  ) |>
  emit()

## `results = "asis"` prints whatever the chunk returns, and gc() returns its summary matrix.
rm(va, vb, ch)
invisible(gc(verbose = FALSE))
```

The two largest movements are in opposite directions and comparable in size.
Land the delivered run scored as ordinary forest is best-quality habitat in the corrected one, and
a nearly equal area the delivered run scored as forest patch is not in a patch at all.
The issue was not a uniform over- or under-count that a single correction factor could absorb; it
misplaced habitat in both directions.

## Radii {#sec-radii}

The Omniscape `radius`, `block_size` and `buffer` are all in **pixels**, not map units.
Radii are derived from quantiles of the interpatch distance distribution, so the corrected patches
imply new radii and hence new run names.

**Both radii move**, and by more than a pixel.

The delivered values are recoverable independently of any stored quantile, because
`write_omniscape_config()` stamps the radius into the run directory name: the `2026-01-23` vintage
wrote `p30_r1507_bs151` (regional) and `p30_r58_bs7` / `p90_r20_bs3` (local).

```{r radii-setup}
ok_q <- is_rebuilt("quantiles_nn_dists") && is_rebuilt("quantiles_all_dists")
qn <- try(tar_read_raw("quantiles_nn_dists"), silent = TRUE)
qa <- try(tar_read_raw("quantiles_all_dists"), silent = TRUE)
ok_q <- ok_q && !inherits(qn, "try-error") && !inherits(qa, "try-error")

if (ok_q) {
  ## `use_radius()` takes the quantile in METRES and returns pixels, so the pin is resolution-
  ## independent: the same distance gives a different pixel count at 30 m and at 90 m.
  px <- function(m, res) ceiling(m / res)

  nn100 <- as.numeric(qn[["100%"]])
  all25 <- as.numeric(qa[["25%"]])

  ## delivered metres, back-derived from the run names above (radius px * pixel size)
  radii_tbl <- data.frame(
    Quantity = c("Regional (all-pairs q25)", "Local (nearest-neighbour q100)"),
    `Delivered m` = c(45202, 1726),
    `Corrected m` = c(all25, nn100),
    `px at 30 m` = c(sprintf("1507 -> %d", px(all25, 30)), sprintf("58 -> %d", px(nn100, 30))),
    `px at 90 m` = c(sprintf("503 -> %d", px(all25, 90)), sprintf("20 -> %d", px(nn100, 90))),
    check.names = FALSE
  )
  radii_tbl$`Delivered m` <- num(radii_tbl$`Delivered m`, 0)
  radii_tbl$`Corrected m` <- num(radii_tbl$`Corrected m`, 0)

  ## The quantile labels are the row names of the vector, and `row.names = FALSE` would throw them
  ## away, leaving a column of distances against nothing. Promote them to a column instead.
  ##
  ## Only one quantile per distribution sets a radius, so that row is bolded rather than given a
  ## column of its own. A delivered-against-corrected column is not possible here: the delivered
  ## store was kept as metadata only, with no objects, so the delivered distributions no longer
  ## exist. The two values that do survive are back-derived from the run directory names and are
  ## in the radii table above.
  quantile_tbl <- function(x, radius_q) {
    d <- data.frame(
      Quantile = names(x),
      `Distance (m)` = num(as.numeric(x), 3),
      check.names = FALSE
    )
    ## Show round quantiles, chosen by name rather than by position.
    ##
    ## The nearest-neighbour distribution is stored in 1% steps. Taking twenty evenly spaced
    ## positions out of those 101 values gave 11%, 32%, 58% and so on: arbitrary numbers that read
    ## as though they had been picked for a reason. Selecting every fifth percentile by name gives
    ## the same coverage and says what it is. A distribution stored more coarsely, as the all-pairs
    ## one is, has no 5% steps to select and is left as it stands.
    wanted <- union(paste0(seq(0, 100, by = 5), "%"), radius_q)
    if (sum(d$Quantile %in% wanted) >= 5) {
      d <- d[d$Quantile %in% wanted, , drop = FALSE]
    }
    attr(d, "bold") <- which(d$Quantile == radius_q)
    d
  }

  nn_tbl <- quantile_tbl(qn, "100%")
  all_tbl <- quantile_tbl(qa, "25%")
}
```

```{r}
#| label: tbl-radii
#| tbl-cap: "Radii before and after the corrections."
#| eval: !expr ok_q
#| results: asis
emit(styled(radii_tbl, booktabs = TRUE, row.names = FALSE))
```

```{r}
#| label: tbl-quantiles-nn
#| tbl-cap: "Nearest-neighbour interpatch distance quantiles, corrected. The bold row is the one the local radius is derived from."
#| eval: !expr ok_q
#| results: asis
emit(styled(
  nn_tbl,
  booktabs = TRUE,
  row.names = FALSE,
  bold_rows = if (length(attr(nn_tbl, "bold"))) attr(nn_tbl, "bold") else NULL
))
```

```{r}
#| label: tbl-quantiles-all
#| tbl-cap: "All-pairs interpatch distance quantiles, corrected. The bold row is the one the regional radius is derived from."
#| eval: !expr ok_q
#| results: asis
emit(styled(
  all_tbl,
  booktabs = TRUE,
  row.names = FALSE,
  bold_rows = if (length(attr(all_tbl, "bold"))) attr(all_tbl, "bold") else NULL
))
```

```{r}
#| label: radii-pending
#| eval: !expr isFALSE(ok_q)
#| results: asis
## `isFALSE(ok_q)` rather than `!ok_q`: a YAML tag cannot be followed by `!`, and `!expr !ok_q`
## reads as two tags in a row.
cat(
  "_Distance quantiles had not been rebuilt when this was rendered;",
  "the radii for the corrected runs are therefore not yet known._\n"
)
```

The regional radius falls by 5.2% and the local one rises by 49%.
Neither is within the roughly one-pixel slack that would leave the pixel count unchanged, so every
corrected run needs its own configuration, which is what the four `2026-08-26` directories are.

The direction differs because the two quantiles measure different things.
The all-pairs q25 fell because the corrected patch set covers less area overall; the
nearest-neighbour q100 is a **maximum**, and maxima are sensitive to the patch-construction issue
in a way that interior quantiles are not.

This is entirely a consequence of the corrections, not of tiling: the serial and tiled builds give
q100 identical to the millimetre ([Appendix B](#sec-tiling-interior)).

## What changed in the outputs {#sec-outputs-changed}

```{r omniscape-outputs-setup}
## Delivered run -> corrected run, matched by scale and resolution. The radii differ because
## they are derived from the corrected interpatch distances; see "Radii".
RUN_PAIRS <- list(
  list(
    scale = "Regional",
    res = 30,
    pub = "2026-01-23_p30_r1507_bs151",
    cor = "2026-08-26_p30_r1429_bs143"
  ),
  list(
    scale = "Regional",
    res = 90,
    pub = "2026-01-23_p90_r503_bs51",
    cor = "2026-08-26_p90_r477_bs49"
  ),
  list(scale = "Local", res = 30, pub = "2026-01-23_p30_r58_bs7", cor = "2026-08-26_p30_r86_bs9"),
  list(scale = "Local", res = 90, pub = "2026-01-23_p90_r20_bs3", cor = "2026-08-26_p90_r29_bs3")
)
OUT_LAYERS <- c("cum_currmap", "flow_potential", "normalized_cum_currmap")

run_tif <- function(run, layer) file.path(QUESNEL_OUT, run, paste0(layer, ".tif"))

## As for the inputs: prose below quotes these inline, so they need to exist even when the
## comparison is skipped.
nz_stat <- stats::setNames(
  rep(NA_real_, 11),
  c(
    "mean_lo",
    "mean_hi",
    "r_lo",
    "r_hi",
    "w5_lo",
    "w5_hi",
    "ha_lo",
    "ha_hi",
    "pct_lo",
    "pct_hi",
    "chan_max"
  )
)
raw_gain <- NA_real_

## Radius is stamped into the run directory name, so the change is recoverable from the names
## themselves rather than having to be restated in the prose and kept in step by hand.
radius_of <- function(run) as.numeric(sub(".*_r([0-9]+)_.*", "\\1", run))
local_radius_gain <- min(vapply(
  Filter(function(p) p$scale == "Local", RUN_PAIRS),
  function(p) 100 * (radius_of(p$cor) / radius_of(p$pub) - 1),
  numeric(1)
))

ok_runs <- all(vapply(
  RUN_PAIRS,
  function(p) {
    all(file.exists(vapply(
      OUT_LAYERS,
      function(l) c(run_tif(p$pub, l), run_tif(p$cor, l)),
      character(2)
    )))
  },
  logical(1)
))

## As for the inputs: one read into a plain vector per raster, then arithmetic in R. Twelve pairs
## at up to 69.2 M cells is 6.8 GB of I/O however it is done, but this way it is done once.
cmp_run <- function(p, layer) {
  ra <- terra::rast(run_tif(p$pub, layer))
  rb <- terra::rast(run_tif(p$cor, layer))
  if (!terra::compareGeom(ra, rb, stopOnError = FALSE)) {
    return(NULL)
  }
  a <- terra::values(ra, mat = FALSE)
  b <- terra::values(rb, mat = FALSE)
  ok <- !is.na(a) & !is.na(b)
  a <- a[ok]
  b <- b[ok]
  d <- b - a
  cellha <- prod(terra::res(ra)) / 1e4
  out <- data.frame(
    scale = p$scale,
    res = p$res,
    layer = layer,
    mean_pub = mean(a),
    mean_cor = mean(b),
    r = suppressWarnings(stats::cor(a, b)),
    ## share of cells whose value moved by no more than 5% -- a cell-level measure of agreement
    ## that, unlike the correlation, is not dominated by the long right tail of current values
    within5 = 100 * mean(abs(d) <= 0.05 * pmax(abs(a), 1e-12)),
    stringsAsFactors = FALSE
  )
  if (layer == "normalized_cum_currmap") {
    ## 1.0 is the only non-arbitrary breakpoint in a normalized current map: it is exactly the
    ## point where current equals flow potential. Above it a cell is channelized, below impeded,
    ## so a cell that crosses it changes its interpretation and not just its value.
    out$ha_cross <- sum((a > 1) != (b > 1)) * cellha
    out$pct_cross <- 100 * mean((a > 1) != (b > 1))
    out$chan_pub <- 100 * mean(a > 1)
    out$chan_cor <- 100 * mean(b > 1)
  } else {
    out$ha_cross <- NA
    out$pct_cross <- NA
    out$chan_pub <- NA
    out$chan_cor <- NA
  }
  rm(a, b, d)
  gc(verbose = FALSE)
  out
}
```

The delivered and corrected runs differ in **two** ways at once, and what is on disk cannot
separate them.
The inputs changed, and the search radii derived from those inputs changed with them.
Both follow from the same fix, so the figures below answer the question "what does the correction do
to the delivered maps".
They do not say how much of that difference comes from the corrected inputs alone.
Separating the two would take a further run, using the corrected inputs at the delivered radius, and
that run has not been made.

This matters most for the two raw current layers.
Cumulative current and flow potential both grow with the size of the search window, and the local
radius grew by at least `r num(local_radius_gain, 0)`%.
Much of the change in those two layers is therefore the larger window rather than the landscape.
Normalized current divides one by the other, which removes most of that effect, and it is the layer
the project's own summaries are built on.
**Normalized current is the layer to read this comparison from.**


```{r}
#| label: tbl-omniscape-outputs
#| tbl-cap: "Omniscape output before and after the corrections, over cells present in both runs. The raw current and flow potential also carry the change in radius; the normalized current largely does not."
#| eval: !expr ok_runs
#| results: asis
outs <- do.call(
  rbind,
  unlist(
    lapply(RUN_PAIRS, function(p) lapply(OUT_LAYERS, cmp_run, p = p)),
    recursive = FALSE
  )
)

pretty_layer <- c(
  cum_currmap = "Cumulative current",
  flow_potential = "Flow potential",
  normalized_cum_currmap = "Normalized current"
)

## Each run contributes three rows, so shading alternates by run rather than by row: the block a
## row belongs to is what the reader needs to see. The normalized current is the layer the
## comparison actually rests on, the other two carrying the change in radius as well, so it is set
## in bold.
run_label <- paste0(outs$scale, ", ", outs$res, " m")

data.frame(
  Run = run_label,
  Layer = unname(pretty_layer[outs$layer]),
  `Mean, delivered` = num(outs$mean_pub, 3),
  `Mean, corrected` = num(outs$mean_cor, 3),
  `Change` = paste0(
    ifelse(outs$mean_cor >= outs$mean_pub, "+", ""),
    num(100 * (outs$mean_cor - outs$mean_pub) / outs$mean_pub, 1),
    " %"
  ),
  `Correlation` = num(outs$r, 3),
  `Cells within 5%` = paste0(num(outs$within5, 1), " %"),
  check.names = FALSE
) |>
  styled(
    stripe_by = run_label,
    bold_rows = which(outs$layer == "normalized_cum_currmap"),
    booktabs = TRUE,
    row.names = FALSE
  ) |>
  emit()

## Quoted inline in the prose that follows, so that the text cannot drift away from the table.
nzr <- subset(outs, layer == "normalized_cum_currmap")
dpct <- 100 * (nzr$mean_cor - nzr$mean_pub) / nzr$mean_pub
nz_stat[] <- c(
  min(dpct),
  max(dpct),
  min(nzr$r),
  max(nzr$r),
  min(nzr$within5),
  max(nzr$within5),
  min(nzr$ha_cross),
  max(nzr$ha_cross),
  min(nzr$pct_cross),
  max(nzr$pct_cross),
  max(abs(nzr$chan_cor - nzr$chan_pub))
)
raw_gain <- min(
  100 *
    (subset(outs, scale == "Local" & layer != "normalized_cum_currmap")$mean_cor /
      subset(outs, scale == "Local" & layer != "normalized_cum_currmap")$mean_pub -
      1)
)
```

```{r omniscape-outputs-pending, eval = !ok_runs, results = "asis"}
cat("_Not every delivered and corrected run pair was on disk when this was rendered._\n")
```

This is seen in the results of the local runs, where cumulative current and flow potential both
rise by at least `r num(raw_gain, 0)`%, tracking the radius rather than anything about the
landscape.
Read down the normalized-current rows instead.

The mean of the normalized current map moves by between `r num(nz_stat[["mean_lo"]], 1)`% and
`r num(nz_stat[["mean_hi"]], 1)`%.
Taken alone that reads as a negligible correction.
It is not, and the next two columns say why.
Agreement between the delivered and corrected maps, measured location by location, is only
`r num(nz_stat[["r_lo"]], 2)` to `r num(nz_stat[["r_hi"]], 2)`, and just
`r num(nz_stat[["w5_lo"]], 0)`% to `r num(nz_stat[["w5_hi"]], 0)`% of the study area is within 5% of
its delivered value.
**The district-wide average is nearly unchanged while the map underneath it is substantially
rearranged.**
That is the same pattern as the issues themselves: a plausible-looking number standing in front of a
wrong one.

```{r}
#| label: tbl-omniscape-crossings
#| tbl-cap: "Cells crossing the normalized-current value of 1, where current equals flow potential. A cell above 1 is channelized, below it impeded, so crossing changes the interpretation of that location and not only its value."
#| eval: !expr ok_runs
#| results: asis
nz <- subset(outs, layer == "normalized_cum_currmap")
data.frame(
  Run = paste0(nz$scale, ", ", nz$res, " m"),
  `Channelized, delivered` = paste0(num(nz$chan_pub, 1), " %"),
  `Channelized, corrected` = paste0(num(nz$chan_cor, 1), " %"),
  `Area crossing (ha)` = num(nz$ha_cross, 0),
  `Share of study area` = paste0(num(nz$pct_cross, 1), " %"),
  check.names = FALSE
) |>
  styled(
    booktabs = TRUE,
    row.names = FALSE
  ) |>
  emit()
```

Between `r num(nz_stat[["ha_lo"]], 0)` and `r num(nz_stat[["ha_hi"]], 0)` ha changes side of that line, which is `r num(nz_stat[["pct_lo"]], 1)`% to `r num(nz_stat[["pct_hi"]], 1)`% of the study area.
The share of the district that is channelized barely moves, by at most
`r num(nz_stat[["chan_max"]], 1)` percentage points.
**The correction moves connectivity around rather than raising or lowering it: the corrected run
finds about as much channelized land as the delivered one did, in different places.**

### Zonal summaries {#sec-zonal-summaries}

`zonal_summaries()` reduces the normalized current map to a mean per biodiversity feature class and
per Landscape Unit x BEC zone combination.
Those means, not the rasters, are what the reporting was written from, so they are the level at
which the correction actually reaches a conclusion.

```{r omniscape-zonal-setup}
zonal_of <- function(run) {
  f <- file.path(QUESNEL_OUT, run, "zonal_by_polys.csv")
  if (!file.exists(f)) {
    return(NULL)
  }
  z <- utils::read.csv(f, stringsAsFactors = FALSE)
  stats::setNames(z$mean_norm_cum_curr, z$zone)
}

zz <- do.call(
  rbind,
  lapply(RUN_PAIRS, function(p) {
    a <- zonal_of(p$pub)
    b <- zonal_of(p$cor)
    if (is.null(a) || is.null(b)) {
      return(NULL)
    }
    k <- intersect(names(a), names(b))
    data.frame(
      run = paste0(p$scale, ", ", p$res, " m"),
      zone = k,
      pub = unname(a[k]),
      cor = unname(b[k]),
      stringsAsFactors = FALSE
    )
  })
)

ok_zonal <- !is.null(zz) && nrow(zz) > 0
if (ok_zonal) {
  ## A handful of high-alpine LU x BEC combinations carry no current at all at the local radius,
  ## in both runs. 0 -> 0 is not a 0% change, it is no basis for a percentage, so those zones are
  ## dropped and counted rather than allowed to propagate a NaN into a median.
  zz$dpct <- ifelse(zz$pub == 0, NA_real_, 100 * (zz$cor - zz$pub) / zz$pub)
}
n_dropped <- if (ok_zonal) sum(is.na(zz$dpct) & grepl("^LU_", zz$zone)) else NA

## The caption for the Landscape Unit table is built here rather than beside the table, because
## knitr evaluates a chunk's options before it evaluates its body: `#| tbl-cap: !expr lu_cap`
## cannot see anything the chunk itself assigns.
lu_cap <- paste(
  "Change in mean normalized current across the Landscape Unit by BEC zone summaries.",
  "The median zone barely moves; the spread around it is large."
)
## Say so in the caption rather than silently showing a smaller denominator than the reader expects.
if (isTRUE(n_dropped > 0)) {
  lu_cap <- paste(
    lu_cap,
    sprintf(
      paste(
        "%d zone-and-run combinations are excluded: a few high-alpine units carry no current at",
        "all at the local radius, in both runs, and 0 to 0 is no basis for a percentage."
      ),
      n_dropped
    )
  )
}
```

```{r}
#| label: tbl-omniscape-zonal-features
#| tbl-cap: "Mean normalized current within each biodiversity feature class, regional run at 30 m. A value above 1 means current concentrates there relative to flow potential."
#| eval: !expr ok_zonal
#| results: asis
FEATURES <- c(
  parks = "Parks",
  ogma = "OGMA",
  wha = "Wildlife habitat areas",
  mdwr = "Mule deer winter range",
  hvmw = "High-value moose wetlands",
  wetlands = "Wetlands"
)
fz <- subset(zz, zone %in% names(FEATURES) & run == "Regional, 30 m")
fz <- fz[order(-fz$dpct), ]

data.frame(
  Feature = unname(FEATURES[fz$zone]),
  `Delivered` = num(fz$pub, 3),
  `Corrected` = num(fz$cor, 3),
  `Change` = paste0(ifelse(fz$dpct >= 0, "+", ""), num(fz$dpct, 1), " %"),
  check.names = FALSE
) |>
  styled(
    booktabs = TRUE,
    row.names = FALSE
  ) |>
  emit()
```

The features do not move together.
Parks, wildlife habitat areas, OGMA and mule deer winter range all gain; the two wetland classes
lose.
The complements tell the same story from the other side: every `non-` zone rises by between 1.4%
and 3.6%, so the corrected run puts relatively more current outside the designated features than
the delivered run did, except in parks, where the gain inside is large enough to run the other way.

Parks is the one to note.
A `r num(abs(subset(zz, zone == "parks" & run == "Regional, 30 m")$dpct), 1)`% change in the mean
normalized current of the protected-areas network is not a rounding correction, and it is the kind
of number that gets quoted on its own.

```{r}
#| label: tbl-omniscape-zonal-lu
#| tbl-cap: !expr lu_cap
#| eval: !expr ok_zonal
#| results: asis
lz <- subset(zz, grepl("^LU_", zone) & !is.na(dpct))
agg <- do.call(
  rbind,
  lapply(split(lz, lz$run), function(g) {
    data.frame(
      Run = g$run[1],
      `Zones` = nrow(g),
      `Median change` = paste0(
        ifelse(stats::median(g$dpct) >= 0, "+", ""),
        num(stats::median(g$dpct), 1),
        " %"
      ),
      `Median absolute` = paste0(num(stats::median(abs(g$dpct)), 1), " %"),
      `90th pct absolute` = paste0(num(stats::quantile(abs(g$dpct), 0.9), 0), " %"),
      `Zones moving > 10%` = num(sum(abs(g$dpct) > 10), 0),
      check.names = FALSE
    )
  })
)
agg <- agg[match(vapply(RUN_PAIRS, function(p) paste0(p$scale, ", ", p$res, " m"), ""), agg$Run), ]

emit(styled(agg, booktabs = TRUE, row.names = FALSE))
```

```{r}
#| label: tbl-omniscape-zonal-extremes
#| tbl-cap: "The six Landscape Unit by BEC zone summaries that move most, regional run at 30 m."
#| eval: !expr ok_zonal
#| results: asis
ez <- subset(zz, grepl("^LU_", zone) & run == "Regional, 30 m" & !is.na(dpct))
ez <- ez[order(-abs(ez$dpct)), ][1:6, ]

## Zone keys are built as LU_<unit name>_<BEC zone>, and unit names themselves contain spaces,
## so split on the LAST underscore. Left as-is they also reach LaTeX full of escaped underscores.
data.frame(
  `Landscape unit and BEC zone` = sub("^LU_(.*)_([^_]+)$", "\\1 (\\2)", ez$zone),
  `Delivered` = num(ez$pub, 2),
  `Corrected` = num(ez$cor, 2),
  `Change` = paste0(ifelse(ez$dpct >= 0, "+", ""), num(ez$dpct, 0), " %"),
  check.names = FALSE
) |>
  styled(
    booktabs = TRUE,
    row.names = FALSE
  ) |>
  emit()
```

These are not small denominators inflating a percentage: the delivered means involved are between
0.8 and 3.1, comfortably in range.
They are zones whose character reverses.
A landscape unit the delivered run reported as one of the strongest current concentrations in the
district comes back as one of the weakest, and another makes the opposite journey.

That is the concrete form of the redistribution the raster comparison measured, and it is the
reason the corrected runs cannot be treated as a refinement of the delivered ones.
The district-wide totals are close enough that they would not, on their own, have prompted a
rerun; the zone-level conclusions drawn from the delivered maps do not survive it.

# Compute cost {#sec-compute-cost}

## Preparing the inputs {#sec-preparing-inputs}

Times in this section are **CPU hours**: processor time as recorded by `targets`, summed over all
branches of a target.
Much of the pipeline runs branches concurrently across workers, so CPU hours are neither wall clock
nor the time a single-threaded run would take.
A stage costing 8 CPU hours across 8 workers occupies about an hour of the day.
They are used here because they are the only figure that compares like with like: the two runs were
made on different machines, under different loads, with different worker counts.
They are grouped by workflow stage rather than by target: the rewrite split, merged and removed
targets, so individual names do not line up between the two runs, and several stages now do in one
target what previously took three.

Two things have to be filtered out or the comparison is meaningless.
The store keeps metadata for targets that were *not rebuilt*, which would compare the delivered run
against itself; and it also keeps metadata for targets the rewrite *removed*, which counts
eliminated work as though it still ran.
Totals below are filtered against `tar_manifest()`, the targets that exist in the pipeline today.
Removed targets are attributed to the stage they used to serve, so each row shows the net effect of the change rather than a deletion and an addition side by side.

```{r}
#| label: tbl-runtime-stage
#| tbl-cap: "Pipeline cost by workflow stage, delivered against corrected (CPU hours)."
#| results: asis
## Group targets into workflow stages. Order matters: first match wins.
stage_of <- function(x) {
  dplyr::case_when(
    grepl("^omniscape", x) ~ "Omniscape configuration",
    grepl("interpatch_distances|quantiles_|forest_patches_chunks|patches_matold", x) ~
      "Interpatch distances",
    grepl("^(resistance|sourcewt)_", x) ~ "Resistance and source-weight surfaces",
    grepl("^patches_|^patch_summary|^forest_patches_final", x) ~ "Patch construction",
    grepl("^forest_disturbance_seral", x) ~ "Seral classification",
    grepl("^VRI_BECNDT", x) ~ "VRI x NDT-BEC overlay",
    grepl("^(readme|reproducibility_receipt)$", x) ~ "Reporting",
    TRUE ~ "Input layers and study area"
  )
}

roll <- function(store) {
  m <- tar_meta(store = store, fields = c("name", "seconds", "parent"))
  br <- m |>
    filter(!is.na(parent)) |>
    group_by(name = parent) |>
    summarise(seconds = sum(seconds, na.rm = TRUE), .groups = "drop")
  tp <- m |> filter(is.na(parent), !is.na(seconds)) |> transmute(name, seconds)
  bind_rows(br, tp) |> group_by(name) |> summarise(seconds = max(seconds), .groups = "drop")
}

if (dir.exists(DELIVERED_META)) {
  ## "The targets that exist in the pipeline today" was a bare `tar_manifest()` when this report
  ## was written, because there was one pipeline and it was the default. That pipeline is retired
  ## and `_targets.R` refuses to run, so the call has to name the projects Quesnel builds through
  ## now. Both of them: the retired script defined the Omniscape configuration and the reporting
  ## targets alongside the dataprep ones, and those stages appear in the table below.
  live <- unlist(
    lapply(
      c("_targets_dataprep_quesnel", "_targets_omniscape_quesnel"),
      function(p) {
        withr::with_envvar(
          c(TAR_PROJECT = p),
          tar_manifest(script = tar_config_get("script", project = p))$name
        )
      }
    ),
    use.names = FALSE
  )
  o <- roll(DELIVERED_META) |> rename(pub = seconds)
  n <- roll(file.path(proj, "_targets")) |> filter(name %in% live) |> rename(cor = seconds)

  by_stage <- full_join(o, n, by = "name") |>
    mutate(stage = stage_of(name)) |>
    group_by(stage) |>
    summarise(
      pub = sum(pub, na.rm = TRUE) / 3600,
      cor = sum(cor, na.rm = TRUE) / 3600,
      .groups = "drop"
    ) |>
    mutate(chg = cor - pub) |>
    arrange(chg)

  tot <- summarise(by_stage, stage = "Total", pub = sum(pub), cor = sum(cor), chg = sum(chg))

  bind_rows(by_stage, tot) |>
    transmute(
      Stage = stage,
      Delivered = sprintf("%.1f h", pub),
      Corrected = sprintf("%.1f h", cor),
      Change = ifelse(abs(chg) < 0.05, "none", sprintf("%+.1f h", chg))
    ) |>
    styled(
      booktabs = TRUE,
      row.names = FALSE,
      align = "lrrr",
      bold_rows = nrow(by_stage) + 1L
    ) |>
    emit()
}
```

**A full build costs about 90 CPU hours less than the delivered one**, despite the corrections
adding work, and nearly all of that comes from two stages.

*Interpatch distances* (-50 h) lost `all_interpatch_distances_chunks`, which held the distances in
memory as a list of chunks; they are now written straight to a parquet dataset.
The quantile step over that dataset was itself rewritten after it was OOM-killed six hours into a
run: asking DuckDB for five quantiles as five aggregates made it keep five sorted states at once and
peak at 336 GB, where one `quantile_cont()` call over a list of probabilities peaks at 28.7 GB and
returns bit-identical values.

*Patch construction* (-40 h) lost `patches_union_final_3` (33.5 h to change 12 polygons out of
162,905, and zero area), and `patches_union_final` itself went from 5 h 52 min to 18 min by tiling.

The one stage that got *slower* is the VRI x NDT-BEC overlay (+0.5 h).
That is the price of correctness: the arcpy LENGTH sliver rule is more expensive than the area rule
it replaced, and it now runs on more and smaller polygons because the seral layer no longer
collapses duplicates.

## Running Omniscape {#sec-running-omniscape}

The saving above is real but it is not the cost of the analysis.
Preparing the inputs is the smaller half of the work; running Omniscape over them is the larger,
and none of the retooling above touches it.

```{r}
#| label: tbl-omniscape-runtime
#| tbl-cap: "Omniscape runs for the corrected inputs, elapsed wall clock. These are the runs themselves, not the preparation of their inputs, and they are unaffected by the retooling described above."
#| results: asis
data.frame(
  Run = c(
    "Regional, 30 m",
    "Regional, 90 m",
    "Local, 30 m",
    "Local, 90 m"
  ),
  Radius = c("1429 px", "477 px", "86 px", "29 px"),
  Workers = c("64", "16", "24", "8"),
  `Elapsed` = c("196 h 44 min", "11 h 08 min", "11 h 28 min", "57 min"),
  check.names = FALSE,
  stringsAsFactors = FALSE
) |>
  styled(
    booktabs = TRUE,
    row.names = FALSE,
    align = "lrrr"
  ) |>
  emit()
```

The regional run at 30 m takes longer on its own than every data-preparation stage in the delivered
and corrected builds put together.
It is the step that sets how long a full analysis takes, and shortening the preparation does not
move it.
What the preparation saving buys is not a faster analysis but a cheaper mistake: a correction that
has to be applied, checked and re-applied costs the preparation time each round, and the Omniscape
runs only once at the end.



{{< pagebreak >}}

# Appendix A: Why the overlay tools were replaced {#sec-appendix-overlay-tools .unnumbered}

## The VRI x NDT-BEC overlay {#sec-vri-becndt .unnumbered}

*Issue 5.*

The leading-group layer is left-joined onto the VRI x NDT-BEC intersection.
This was done with `terra::union()` followed by dropping the leading-group-only parts.
`terra::union()` is wrong here: it emits output polygons that overlap one another, and its dissolved
footprint is smaller than its input, so it double-counts some land and drops other land outright.
The two errors act in opposite directions, which is why the delivered total looked only slightly low
rather than obviously broken.

`terra::intersect()` plus `terra::erase()` partitions the input exactly.

```{r}
#| label: tbl-vri-becndt
#| tbl-cap: "VRI x NDT-BEC overlay, whole study area. In the corrected layer the parts sum exactly to the area they cover."
data.frame(
  Measure = c("features", "sum of parts (ha)", "dissolved footprint (ha)", "double-counted (ha)"),
  Delivered = c("400,534", "3,534,286.2", "3,518,837.1", "15,449.2"),
  Corrected = c("398,109", "3,541,465.4", "3,541,465.4", "0.0"),
  stringsAsFactors = FALSE
) |>
  styled(
    booktabs = TRUE
  )
```

## The `terra::union()` issue {#sec-union-issue .unnumbered}

*Issue 5, and the cause behind it.*

The single largest cause of wrong numbers here was one function, reached for in three separate
places because the project's own notes recommended it: `terra::union()` has exactly the overlay
semantics the arcpy scripts need, and produces silently wrong results on this data.

It fails in two directions at once, which is why the totals looked plausible rather than broken.
Its output polygons overlap one another, so their areas sum to more ground than they cover, and the ground they cover is smaller than the inputs, so land goes missing.
The two errors partly cancel.

```{r}
#| label: tbl-union-defect
#| tbl-cap: "The same issue measured at three scales."
data.frame(
  Where = c(
    "VRI x NDT-BEC overlay (study area)",
    "Final resultant (2,933-polygon window)",
    "Minimal case (7 polygons)"
  ),
  `Counted twice` = c("15,449 ha", "303 ha", "5.7 ha"),
  `Land lost` = c("22,628 ha", "82 ha", "1,203 ha"),
  `Slower by` = c("418x", "28.5x", "n/a"),
  check.names = FALSE,
  stringsAsFactors = FALSE
) |>
  styled(booktabs = TRUE)
```

Three things about it are worth recording, because each one delayed finding it.

It is not a version problem.
The identical numbers come out under terra 1.8.86 and 1.9.34, on both R 4.5.3 and R 4.6.1, with GDAL 3.8.4 and GEOS 3.12.1, so it is not fixed in the current release.
It had not been reported, and is now filed as [rspatial/terra#2175](https://github.com/rspatial/terra/issues/2175).

It does not reproduce on synthetic data.
On clean, hand-built polygons `union()` is exact.
Only real inventory geometry triggers it, so a passing test suite is no evidence that a call is safe.

It is a precision failure.
Shifting the vertices by a single millimetre makes the answer come out right, to within 0.14 m² of 12.3 million:

```{r}
#| label: tbl-union-precision
#| tbl-cap: "Simplifying the inputs before the union, on the minimal case. A millimetre is far below the precision of the source data."
data.frame(
  Tolerance = c("none", "1 mm", "1 cm", "10 cm", "1 m", "10 m"),
  `Counted twice (m2)` = c("56,640.5", "0.0", "0.0", "0.0", "0.0", "299.1"),
  `Error vs truth (m2)` = c("-12,030,541.2", "-0.1", "+0.3", "+6.9", "+1,222.6", "+8,570.4"),
  check.names = FALSE,
  stringsAsFactors = FALSE
) |>
  styled(
    booktabs = TRUE
  )
```

That last point offers a tempting workaround, and the pipeline deliberately does not use it.
The replacement, `overlay_left_join()`, is `intersect()` plus `erase()`.
It partitions the input exactly by construction rather than by luck, it is faster, and it does not depend on a tolerance that happens to suit this data and silently stops working at 10 m.

The seven-polygon case, the figure above and the measurements are attached to [rspatial/terra#2175](https://github.com/rspatial/terra/issues/2175), which was **fixed upstream on 2026-08-29**.
The project now runs terra 1.9-48, installed from GitHub and pinned by commit in `renv.lock`, and the fix is confirmed here: the minimal case goes from 56,640 m² of self-overlap to 0.000, and the study-area window from an 8,607 m² footprint error to 0.012 m².

## A second issue, in `terra::erase()` {#sec-erase-issue .unnumbered}

*Encountered while correcting issue 5.*

Replacing `union()` was not the end of it.
The rewritten step then ran for six hours and died with `[[<-,SpatVector] cannot add these values`, which named neither a column nor a reason.
The cause was `terra::erase()` returning a `SpatVector` with **more attribute rows than geometries**: a polygon whose difference comes out empty loses its geometry but keeps its row.
Nothing reports it at the time, and whatever next reads the attributes fails instead.

It is a different problem from the first, and the fix for #2175 does not cover it: that fix moved `union()` onto `erase_agg()`, which is the function R's `erase()` was already calling.
Reported as [rspatial/terra#2179](https://github.com/rspatial/terra/issues/2179), still open.

```{r}
#| label: tbl-erase-defect
#| tbl-cap: "Erasing with terra against erasing with sf, on the same 42 tiles. sf cannot desynchronise: its attributes are columns of the same data frame as the geometry."
data.frame(
  Measure = c(
    "tiles where the overlay is consistent",
    "total time for the erase step",
    "agreement where terra succeeded"
  ),
  terra = c("37 of 42", "192.5 s", "n/a"),
  sf = c("42 of 42", "388.9 s", "0.000000 m\u00b2"),
  stringsAsFactors = FALSE
) |>
  styled(
    booktabs = TRUE
  )
```

The pipeline therefore uses `sf::st_difference()` for this one step, through `erase_polygons()`.
An earlier version tried terra first and fell back to sf only when the result was inconsistent, and that was dropped deliberately.
It made the output depend on which tiles happened to trip a problem nobody can characterise, and one code path is easier to trust than two.
The measured cost of using sf throughout is about twice the terra time on a step that tiles to roughly five minutes.

Two things about this are worth recording.
It does not reproduce on synthetic geometry either.
Hand-built degenerate polygons come back *invalid*, and terra handles those correctly; the polygon that triggers it is a valid four-vertex ring of zero area whose second and third vertices differ by a ten-billionth of a metre, in effect a line traced out and back.
The failure is also silent by construction.
That is why the check now sits inside `overlay_left_join()`, comparing geometry and attribute counts after every overlay so that a malformed object fails there rather than travelling on.

### Fixed upstream, after this work {#sec-fixed-upstream .unnumbered}

`terra::erase()` was fixed in [`12df7fbd`](https://github.com/rspatial/terra/commit/12df7fbd4f85eb01faa02c580a31ce7e34275c6a).
That fix is **not** in the build this pipeline runs: `renv.lock` pins terra at `1a8843f`, which
carries the #2175 fix only, so the behaviour described above is present in the version that produced these results.
The workaround is therefore still in place and still necessary.

The workaround has since been promoted to `spatialutils` as `erase_polygons()`, together with
`overlay_left_join()` and `keep_polygons()`.
It survives the fix on its own merits: it also tolerates an empty right-hand layer, and it drops the
non-polygonal residue an overlay can leave behind.
Neither of those is a terra issue.

The second of the two is worth recording because it bit during this work.
When everything left of a polygon is a shared edge, `sf::st_difference()` can return a bare `LINESTRING` sitting directly in the geometry column rather than inside a `GEOMETRYCOLLECTION`.
`terra::vect()` then drops that geometry while keeping its attribute row, producing exactly the
desynchronisation #2179 causes, from an entirely different direction.
Measured on one tile of the study area, one of 6,871 differences was such a line.
A guard testing only for `GEOMETRYCOLLECTION` misses it; `spatialutils::intersect_clean()` carried
the same too-narrow guard and has been fixed in the same release.

{{< pagebreak >}}

# Appendix B: Per-target timings and tiling {#sec-appendix-timings .unnumbered}

## Tiling the final resultant {#sec-tiling-resultant .unnumbered}

`patches_union_final` is the arcpy step-6 overlay: the seral layer joined to both interior-forest
layers.
It ran as a single process and took **5 h 51 min 58 s**.

The overlay can be split across space without changing its result, so it now branches over `study_area_tiles`, the same 8x8 grid the seral layer itself was built on.
It reads its base layer through GDAL rather than out of the `targets` store, so a worker materialises only the polygons inside its own tile.

```{r}
#| label: tbl-tiling-cost
#| tbl-cap: "Building the final resultant, serial against tiled."
#| results: asis
data.frame(
  Quantity = c("Runtime", "Branches", "Slowest branch", "Features", "Planar area"),
  Serial = c("5 h 51 min 58 s", "1", "n/a", "189,666", "2,473,003.272 ha"),
  Tiled = c("17 min 48 s", "49, all succeeding", "2 min 48 s", "192,683", "2,473,003.272 ha"),
  check.names = FALSE
) |>
  styled(
    booktabs = TRUE,
    row.names = FALSE
  ) |>
  emit()
```

**19.8x faster, with the area preserved exactly.**

Reusing the *same* grid is load-bearing, not a convenience.
The seral layer already carries vertices lying bit-exactly on these seam coordinates, so re-cutting
on the same grid intersects at existing vertices and GEOS invents no new ones; the halves of a split
polygon abut exactly and dissolve back together.
A finer grid would cut at fresh coordinates and forfeit that.

Cropping splits polygons at the seams, so the resultant carries 3,017 more features (+1.59%).
That is a change to a shipped artefact, `forest_patches_final.gpkg`, even though no hectare moves.
It does not propagate: `calc_matold()` dissolves before the interpatch distances are measured.

The verification that matters is the distance distribution, since that is what sets the Omniscape
radii:

```{r}
#| label: tbl-tiling-quantiles
#| tbl-cap: "Nearest-neighbour distance quantiles in metres, serial against tiled. The 100% row is the one the Omniscape radius is derived from."
#| results: asis
data.frame(
  Quantile = c("25%", "50%", "75%", "90%", "99%", "100%"),
  Serial = c(0.372, 12.500, 39.787, 91.901, 330.123, 2566.803),
  Tiled = c(0.464, 12.603, 39.899, 92.131, 330.685, 2566.803)
) |>
  transform(Difference = sprintf("%+.3f", Tiled - Serial)) |>
  styled(
    booktabs = TRUE,
    row.names = FALSE
  ) |>
  emit()
```

Every quantile agrees to under a metre, and q100, the one the configuration actually consumes, is identical.
The radius is unchanged either way.

Dissolving does move slightly: `patches_matold` comes out 68,255 patches tiled against 68,401
serial.
This is the dissolve, not the tiling.
Before the dissolve the two layers agree to **0.000000 ha**; after it, the *serial* dissolve has
gained 0.798 ha, which a dissolve cannot do.
`terra::aggregate()` is simply not exact on a layer of this size.
The same layers self-overlap by about 10 ha out of 2.47 Mha in both builds.

## Tiling the interior-forest overlay {#sec-tiling-interior .unnumbered}

The final resultant was not the only overlay of this shape.
`patches_interior_forest_old` and `patches_interior_forest_mature_old` erase a buffer mask from the
old and mature+old patch layers, and between them they cost **8.1 CPU hours** in the corrected run: 3 h 07 min and 5 h 00 min respectively.
They were the largest remaining single-target costs once the resultant was tiled.

The cause is structural rather than incidental.
`patches_create_erase_mask()` rbinds the four buffer layers and dissolves them with
`terra::aggregate()`, which leaves the mask as a **single multipolygon covering 2,401,307.9 ha**, the whole study area.
The cost of a difference is superlinear in the complexity of the layer being subtracted, so each of
the 42,764 patches was differenced against study-area-wide geometry no matter how small the patch
itself was.

Cropping both sides to a tile first shrinks the mask each difference sees to tile size.
It reuses the same 49-tile grid as the resultant, for the same reason given above.

```{r}
#| label: tbl-interior-tiling-cost
#| tbl-cap: "Building the old interior-forest layer, untiled against tiled. Sequential in both cases, so the tiled column is compute, not parallelism."
#| results: asis
data.frame(
  Quantity = c("Runtime", "Tiles", "Slowest tile", "Features", "Planar area"),
  Untiled = c("3 h 07 min 34 s", "1", "n/a", "10,263", "167,496.729062 ha"),
  Tiled = c("6 min 08 s", "49, all succeeding", "1 min 27 s", "10,263", "167,496.729059 ha"),
  check.names = FALSE
) |>
  styled(
    booktabs = TRUE,
    row.names = FALSE
  ) |>
  emit()
```

**30.5x faster, with the same 10,263 features and the area agreeing to three parts in a hundred billion**, which is floating-point noise rather than a difference.
Unlike the resultant, the feature count does not change: the layer is dissolved after the tiles are
combined rather than per tile, so halves split at a seam are rejoined before anything downstream
sees them.

Tiling is only valid because this overlay is *local*: which part of a patch survives depends solely
on mask geometry near that patch.
The same treatment would be wrong for the dissolve or the interpatch distances, where a feature's
result depends on the whole layer.

The per-tile spread is the useful diagnostic: 0.2 s at the fastest, 2.8 s median, 87.5 s at the
slowest.
A factor of 31 between median and worst says cost is driven by local mask complexity rather than by
patch count, which is what makes the untiled form scale so badly: every patch pays the worst case.

### Why this matters beyond Quesnel {#sec-beyond-quesnel .unnumbered}

Quesnel is not the largest of the Cariboo districts this workflow is intended to cover, nor the
smallest: 100 Mile House is smaller, and Cariboo-Chilcotin is very much larger.
Cariboo-Chilcotin is roughly **2.4 times** Quesnel's area, and the costs above do not scale with area: the mask for a larger district is both larger and more intricate, and the superlinearity is in mask complexity, not in patch count.
At the untiled cost profile, the interior-forest pair alone would plausibly dominate an entire
district build, and it is not the only overlay of this shape in the workflow.

This retooling is therefore a **prerequisite** for extending the analysis beyond Quesnel rather
than an optimisation of it.
The arithmetic is unforgiving in the other direction too: a step that takes a day is not merely slow,
it is a step that cannot be retried casually when something downstream needs changing, and this
report exists because several things did need changing.

One caveat on reading the runtime figures elsewhere in this report against this section.
Quesnel's stored results were built before this overlay was tiled, so the costs reported there are
what that build actually took.
The figures above are measured on the same inputs and reproduce the stored output exactly, but they
are not what produced the numbers elsewhere in this document.
Districts built since carry the tiled form from the outset.

## Memory {#sec-memory .unnumbered}

Measured by sampling the resident set size of every project R process every 30 seconds across the
rerun, rather than estimated.

```{r}
#| label: tbl-memory
#| tbl-cap: "Memory use of the data-preparation pipeline with 8 workers. Omniscape is excluded: it is opt-in and needs hundreds of GB."
data.frame(
  Measure = c(
    "Peak aggregate RSS",
    "Largest single process",
    "Typical during tiled overlays",
    "Per-worker ceiling for raster operations"
  ),
  Value = c("110.0 GB", "41.1 GB", "~1.5 GB per branch", "~10 GB"),
  stringsAsFactors = FALSE
) |>
  styled(
    booktabs = TRUE
  )
```

## Run naming {#sec-run-naming .unnumbered}

Omniscape run directories are stamped with `OMNISCAPE_VINTAGE`, set in `_targets.R`.
It was previously a hardcoded `"2026-01-23"`, which is the vintage of the delivered runs.
Because `run_omniscape()` deletes an existing output directory before running, a corrected run whose
radius matched a delivered one would have destroyed it, and there are 17 GB of delivered results,
four of them under that prefix, including the run the report was written from.
Corrected runs are stamped `2026-08-26`.

{{< pagebreak >}}

# Appendix C: Reproducing this report {#sec-appendix-reproducing .unnumbered}

This report is deliberately **not** a pipeline target: it compares two runs, one of which no longer
exists as a live `targets` store.
It reads:

- `published-2026-01-27/`, the delivered patches layer, composite rasters and map,
  copied aside before the rerun overwrote them;
- the `targets` metadata snapshot of the delivered run;
- the live `_targets` store for the corrected values.

Render with:

```bash
QUARTO_R=/path/to/R quarto render reports/corrections-2026-08.qmd \
  --to pdf --output-dir ../Outputs
```

The rendered PDF is written to `Outputs/`, not next to the source, so it lands on the shared volume
alongside the other results rather than inside the clone.
`QUARTO_R` pins the render to the R version `renv.lock` names rather than whichever R is first on
the path, since the project library is per R minor version.

The output file is named in the document's front matter rather than on the command line.
It carries a `-cariboo` suffix because `Outputs/` is shared storage rather than per-branch, and the
Quesnel branch renders a report of the same name; without distinct names the two overwrite each
other depending on which was rendered last.

```{r}
#| label: tbl-session
#| tbl-cap: "Environment."
## GEOS and GDAL come from sf rather than being read separately: they are the versions sf is
## actually linked against, which is what determines the overlay behaviour this report describes.
geos_ver <- sub(".*GEOS ([0-9.]+).*", "\\1", sf::sf_extSoftVersion()[["GEOS"]])
gdal_ver <- sf::sf_extSoftVersion()[["GDAL"]]

data.frame(
  Item = c(
    "Operating system", "Platform", "R",
    "targets", "terra", "sf", "GEOS", "GDAL",
    "git branch", "git commit"
  ),
  Value = c(
    utils::osVersion,
    R.version$platform,
    as.character(getRversion()),
    as.character(packageVersion("targets")),
    as.character(packageVersion("terra")),
    as.character(packageVersion("sf")),
    geos_ver,
    gdal_ver,
    system("git rev-parse --abbrev-ref HEAD", intern = TRUE),
    system("git rev-parse --short HEAD", intern = TRUE)
  ),
  stringsAsFactors = FALSE
) |>
  styled(booktabs = TRUE)
```
