---
title: "Quesnel NRD connectivity: corrections and rerun, August 2026"
subtitle: "What changed, what it cost, and what it means for the Omniscape runs"
author: "Alex Chubaty"
date: "Last updated: `r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: pdflatex
geometry: margin=2.5cm
urlcolor: blue
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center",
  ## Keep figures in the section that discusses them. Captioned figures are LaTeX floats,
  ## so by default two full-width 4800 px maps drift to wherever they fit -- which put them
  ## pages away from the text about them. `H` (from the float package, loaded in the YAML
  ## header) pins them exactly where they appear in the source.
  fig.pos = "H", out.extra = ""
)

## knitr evaluates chunks in the directory of the input file, so anything that resolves the project
## by falling back to the working directory (`workflowtools::findProjectPath()` among them) reports
## `reports/` and every derived path silently misses. Walk up to the directory holding `_targets.R`.
find_project_root <- function(start = getwd()) {
  p <- normalizePath(start, mustWork = FALSE)
  while (!file.exists(file.path(p, "_targets.R")) && dirname(p) != p) p <- dirname(p)
  if (!file.exists(file.path(p, "_targets.R"))) stop("could not locate the project root")
  p
}

proj <- find_project_root()
knitr::opts_knit$set(root.dir = proj)

## Emit native LaTeX tables rather than markdown ones. Pandoc turns a markdown table into a
## `longtable`, which splits across pages -- so a table lands half on one page and half on the next.
## `format = "latex"` with `booktabs` produces a `table` float instead, and floats never split; the
## preamble then pins them in place and sets them a size down.
options(knitr.table.format = "latex")

library(dplyr)
library(knitr)
library(targets)

## Artefacts of the published (pre-correction) run, copied out of the way before the rerun
## overwrote them. See section "Reproducing this report".
PUBLISHED <- file.path(proj, "Data", "processed", "published-2026-01-27")
## Derived from wherever `_targets` actually resolves to (it is usually a symlink to bulk storage),
## so no site-specific absolute path is baked into the repo. Override with the environment variable.
PUB_META <- Sys.getenv(
  "BC_CONN_PUBLISHED_META",
  file.path(dirname(normalizePath(file.path(proj, "_targets"), mustWork = FALSE)),
            "_targets_meta_backup_pre-2026-08-25-fixes")
)
## Figures live beside this document, wherever it is. Building the path out of `proj` instead
## hardcodes the assumption that this file sits in `reports/`, and when project-root resolution
## slipped that assumption produced a nested `reports/reports/figures`.
rmd_dir <- tryCatch(dirname(knitr::current_input(dir = TRUE)), error = function(e) NULL)

if (is.null(rmd_dir) || !nzchar(rmd_dir) || is.na(rmd_dir)) {
  rmd_dir <- file.path(proj, "reports") ## sourcing the chunks outside knitr
}

FIGDIR <- normalizePath(file.path(rmd_dir, "figures"), mustWork = FALSE)
dir.create(FIGDIR, showWarnings = FALSE, recursive = TRUE)

## One place that decides how every table in this report is drawn.
##
## `HOLD_position` keeps a table with the text discussing it instead of letting LaTeX float it
## elsewhere, and the smaller size keeps the wider ones on a single page. Written with `kableExtra`
## rather than raw LaTeX in the document preamble, so it can be adjusted by anyone comfortable with
## R without having to read TeX.
##
## Captions are plain text on purpose: with native LaTeX tables the caption is passed through
## verbatim, so a stray underscore or backtick becomes a TeX error reported several pages away from
## the caption that caused it.
styled <- function(x, ...) {
  knitr::kable(x, ...) |>
    kableExtra::kable_styling(latex_options = "HOLD_position", font_size = 9)
}

ha  <- function(x) sum(terra::expanse(x, unit = "ha", transform = FALSE))
num <- function(x, d = 1) formatC(round(x, d), format = "f", big.mark = ",", digits = d)
dur <- function(s) {
  ifelse(is.na(s), "--",
    ifelse(s >= 3600, sprintf("%.2f h", s / 3600),
      ifelse(s >= 60, sprintf("%.1f m", s / 60), sprintf("%.1f s", s))))
}
have <- function(...) all(file.exists(...))

## `tar_read()` happily returns a target that has NOT been rebuilt: the store keeps the published
## run's object and metadata until the target is remade. Reading one of those and calling it
## "corrected" would silently compare the published run against itself. A target counts as rebuilt
## only where its recorded time differs from the published snapshot.
secs_by_name <- function(store) {
  if (!dir.exists(store)) return(NULL)
  m <- tar_meta(store = store, fields = c("name", "seconds"))
  stats::setNames(m$seconds, m$name)
}

PUB_SECONDS <- secs_by_name(PUB_META)
CUR_SECONDS <- secs_by_name(file.path(proj, "_targets"))

is_rebuilt <- function(target) {
  if (is.null(PUB_SECONDS) || !target %in% names(PUB_SECONDS)) return(TRUE)
  if (is.null(CUR_SECONDS) || !target %in% names(CUR_SECONDS)) return(FALSE)
  cur <- CUR_SECONDS[[target]]
  if (is.na(cur)) return(FALSE)
  !isTRUE(all.equal(cur, unname(PUB_SECONDS[[target]])))
}
```

# Summary

A defect found in the seral-patch construction (commit `145cffb`) turned out to be one of several,
and together they invalidate the connectivity results published in January 2026.
This report documents what was wrong, what the corrected numbers are, what the rerun cost, and how
urgently the Omniscape runs need to be repeated.

The headline is that the two layers Omniscape actually consumes -- the composite resistance and
source-weight rasters -- change materially, because roughly half of what the published run treated
as old interior forest was not forest at all, and a further large block of genuine old forest was
modelled at the wrong resistance.
The Omniscape runs therefore need to be repeated; they cannot be patched.

A second strand of work, in the runtime section below, addressed the two largest spatial overlays.
Tiling the final resultant is done and took it from 5 h 52 m to 18 m.
The interior-forest overlay is the remaining one: tiling it measures 3 h 08 m down to 6 m on a
prototype, reproducing the stored output exactly, but that change is not in this branch.
That matters here because the corrections below had to be applied, checked and re-applied several
times over, and a step measured in hours is one that cannot be retried casually.
It matters more for what comes next: the workflow is intended to cover the wider Cariboo, and
Cariboo-Chilcotin alone is roughly 2.4 times Quesnel's area, with these costs growing faster than
area rather than in step with it.
The retooling is a prerequisite for that extension rather than a refinement of it.

```{r headline}
data.frame(
  Item = c(
    "Seral layer, total area",
    "Old interior forest",
    "VRI x NDT-BEC layer, double-counted area",
    "Pipeline steps eliminated"
  ),
  Published = c("2,931,558 ha", "326,081.5 ha (13,783 polys)", "15,449.2 ha", "--"),
  Corrected = c("see Section 3", "166,915.6 ha (10,348 polys)", "0.0 ha", "35.1 h of compute"),
  stringsAsFactors = FALSE
) |>
  styled(caption = "Headline effects of the corrections.", booktabs = TRUE)
```

# Defects

Each of these produced plausible-looking wrong numbers rather than an error, which is why they
survived into a published run.

```{r defects}
data.frame(
  Defect = c(
    "Interior forest erased the 25 m buffer from the mature+old target as well as old",
    "Old/mature-old patches were never filtered to INTERIOR_CATEGORY MO/O",
    "Interior forest overwrote `Seral`; the union propagated the wrong attributes",
    "Seral layer built with `sf::st_join()` instead of an intersection",
    "Leading group joined with `terra::union()`",
    "Sliver elimination used the area rule, not arcpy's LENGTH rule",
    "Patch statistics measured overlay fragments",
    "Omniscape tile overlap divided by pixel size"
  ),
  Effect = c(
    "mature+old interior forest came out byte-identical to old-only",
    "50.8% of `old interior forest` was unclassified land",
    "160,417 ha of old cores modelled at resistance 250 instead of 1",
    "layer inflated to 7.78 M polygons / 39.5 Mha",
    "22,628 ha of land dropped, 15,449 ha double-counted",
    "not the semantics the arcpy scripts this ports rely on",
    "smallest reported Old patch was 4.6e-6 m2",
    "artifact-ridden mosaic in the tiled benchmark run"
  ),
  stringsAsFactors = FALSE
) |>
  styled(caption = "Defects corrected in this rerun.", booktabs = TRUE)
```

# Data corrections

## The VRI x NDT-BEC overlay

The leading-group layer is left-joined onto the VRI x NDT-BEC intersection.
This was done with `terra::union()` followed by dropping the leading-group-only parts.
`terra::union()` is wrong here: it emits output polygons that overlap one another, and its dissolved
footprint is smaller than its input, so it double-counts some land and drops other land outright.
The two errors act in opposite directions, which is why the published total looked only slightly low
rather than obviously broken.

`terra::intersect()` plus `terra::erase()` partitions the input exactly.

```{r vri-becndt}
data.frame(
  Measure = c("features", "sum of parts (ha)", "dissolved footprint (ha)", "double-counted (ha)"),
  Published = c("400,534", "3,534,286.2", "3,518,837.1", "15,449.2"),
  Corrected = c("398,109", "3,541,465.4", "3,541,465.4", "0.0"),
  stringsAsFactors = FALSE
) |>
  styled(caption = paste(
    "VRI x NDT-BEC overlay, whole study area.",
    "In the corrected layer the parts sum exactly to the area they cover."
  ), booktabs = TRUE)
```

## The `terra::union()` defect

The single largest cause of wrong numbers here was one function, reached for in three separate
places because the project's own notes recommended it: `terra::union()` has exactly the overlay
semantics the arcpy scripts need, and produces silently wrong results on this data.

It fails in two directions at once, which is why the totals looked plausible rather than broken. Its
output polygons overlap one another, so their areas sum to more ground than they cover; and the
ground they cover is smaller than the inputs, so land goes missing. The two errors partly cancel.

```{r union-defect}
data.frame(
  Where = c("VRI x NDT-BEC overlay (study area)",
            "Final resultant (2,933-polygon window)",
            "Minimal case (7 polygons)"),
  `Counted twice` = c("15,449 ha", "303 ha", "5.7 ha"),
  `Land lost` = c("22,628 ha", "82 ha", "1,203 ha"),
  `Slower by` = c("418x", "28.5x", "--"),
  check.names = FALSE,
  stringsAsFactors = FALSE
) |>
  styled(caption = "The same defect measured at three scales.", booktabs = TRUE)
```

Three things about it are worth recording, because each one delayed finding it.

It is not a version problem. The identical numbers come out under terra 1.8.86 and 1.9.34, on both
R 4.5.3 and R 4.6.1, with GDAL 3.8.4 and GEOS 3.12.1, so it is not fixed in the current release. It
had not been reported, and is now filed as
[rspatial/terra#2175](https://github.com/rspatial/terra/issues/2175).

It does not reproduce on synthetic data. On clean, hand-built polygons `union()` is exact. Only real
inventory geometry triggers it, so a passing test suite is no evidence that a call is safe.

It is a precision failure. Shifting the vertices by a single millimetre makes the answer come out
right, to within 0.14 m² of 12.3 million:

```{r union-precision}
data.frame(
  Tolerance = c("none", "1 mm", "1 cm", "10 cm", "1 m", "10 m"),
  `Counted twice (m2)` = c("56,640.5", "0.0", "0.0", "0.0", "0.0", "299.1"),
  `Error vs truth (m2)` = c("-12,030,541.2", "-0.1", "+0.3", "+6.9", "+1,222.6", "+8,570.4"),
  check.names = FALSE,
  stringsAsFactors = FALSE
) |>
  styled(caption = paste(
    "Simplifying the inputs before the union, on the minimal case.",
    "A millimetre is far below the precision of the source data."
  ), booktabs = TRUE)
```

That last point offers a tempting workaround, and the pipeline deliberately does not use it. The
replacement, `overlay_left_join()`, is `intersect()` plus `erase()`: it partitions the input exactly
by construction rather than by luck, it is faster, and it does not depend on a tolerance that
happens to suit this data and silently stops working at 10 m.

The seven-polygon case, the figure above and the measurements are attached to
[rspatial/terra#2175](https://github.com/rspatial/terra/issues/2175), which was **fixed upstream on
2026-08-29**. The project now runs terra 1.9-48, installed from GitHub and pinned by commit in
`renv.lock`, and the fix is confirmed here: the minimal case goes from 56,640 m² of self-overlap to
0.000, and the study-area window from a 8,607 m² footprint error to 0.012 m².

## A second defect, in `terra::erase()`

Replacing `union()` was not the end of it. The rewritten step then ran for six hours and died with
`[[<-,SpatVector] cannot add these values`, which named neither a column nor a reason. The cause was
`terra::erase()` returning a `SpatVector` with **more attribute rows than geometries**: a polygon
whose difference comes out empty loses its geometry but keeps its row. Nothing reports it at the
time; whatever next reads the attributes fails instead.

It is a different bug from the first, and the fix for #2175 does not cover it: that fix moved
`union()` onto `erase_agg()`, which is the function R's `erase()` was already calling. Reported as
[rspatial/terra#2179](https://github.com/rspatial/terra/issues/2179), still open.

```{r erase-defect}
data.frame(
  Measure = c("tiles where the overlay is consistent", "total time for the erase step",
              "agreement where terra succeeded"),
  terra = c("37 of 42", "192.5 s", "--"),
  sf = c("42 of 42", "388.9 s", "0.000000 m2"),
  stringsAsFactors = FALSE
) |>
  styled(caption = paste(
    "Erasing with terra against erasing with sf, on the same 42 tiles.",
    "sf cannot desynchronise: its attributes are columns of the same data frame as the geometry."
  ), booktabs = TRUE)
```

The pipeline therefore uses `sf::st_difference()` for this one step, through `erase_polygons()`. An
earlier version tried terra first and fell back to sf only when the result was inconsistent; that
was dropped deliberately. It made the output depend on which tiles happened to trip a bug nobody can
characterise, and one code path is easier to trust than two. The measured cost of using sf
throughout is about twice the terra time on a step that tiles to roughly five minutes.

Two things about this defect are worth recording. It does not reproduce on synthetic geometry
either: hand-built degenerate polygons come back *invalid*, and terra handles those correctly. The
polygon that triggers it is a valid four-vertex ring with zero area, whose second and third vertices
differ by 1e-10 -- a line traced out and back. And the failure is silent by construction, which is
why the check now sits inside `overlay_left_join()`: it compares geometry and attribute counts after
every overlay and fails there, rather than letting a malformed object travel.

### Fixed upstream, after this work

`terra::erase()` was fixed in [`12df7fbd`](https://github.com/rspatial/terra/commit/12df7fbd4f85eb01faa02c580a31ce7e34275c6a).
That fix is **not** in the build this pipeline runs: `renv.lock` pins terra at `1a8843f`, which
carries the #2175 fix only, so the defect described above is present in the version that produced
these results -- which is why the workaround is still in place and still necessary.

The workaround has since been promoted to `spatialutils` as `erase_polygons()`, together with
`overlay_left_join()` and `keep_polygons()`.
It survives the fix on its own merits: it also tolerates an empty right-hand layer, and it drops the
non-polygonal residue an overlay can leave behind.
Neither of those is a terra defect.

The second of the two is worth recording because it bit during this work.
`sf::st_difference()` can return a bare `LINESTRING` sitting directly in the geometry column -- not
inside a `GEOMETRYCOLLECTION` -- when everything left of a polygon is a shared edge.
`terra::vect()` then drops that geometry while keeping its attribute row, producing exactly the
desynchronisation #2179 causes, from an entirely different direction.
Measured on one tile of the study area, one of 6,871 differences was such a line.
A guard testing only for `GEOMETRYCOLLECTION` misses it; `spatialutils::intersect_clean()` carried
the same too-narrow guard and has been fixed in the same release.

## Forest patches

```{r patches-table}
pub_gpkg <- file.path(PUBLISHED, "forest_patches_final.gpkg")
ok_patches <- have(pub_gpkg) &&
  !inherits(try(tar_read_raw("forest_patches_final"), silent = TRUE), "try-error") &&
  is_rebuilt("forest_patches_final")
```

```{r patches-compare, eval = ok_patches}
tabulate_patches <- function(v) {
  v$ha <- terra::expanse(v, unit = "ha", transform = FALSE)
  as.data.frame(v) |>
    group_by(Seral, Resistance, SourceWt) |>
    summarise(n = n(), ha = sum(ha), .groups = "drop")
}

b <- terra::vect(pub_gpkg)
a <- tidyterra::as_spatvector(tar_read_raw("forest_patches_final"))

cmp <- full_join(
  tabulate_patches(b) |> rename(n_pub = n, ha_pub = ha),
  tabulate_patches(a) |> rename(n_cor = n, ha_cor = ha),
  by = c("Seral", "Resistance", "SourceWt")
) |>
  mutate(across(where(is.numeric), \(x) ifelse(is.na(x), 0, x)),
         d = ha_cor - ha_pub,
         pct = ifelse(ha_pub > 0, 100 * d / ha_pub, NA_real_)) |>
  arrange(Resistance)

cmp |>
  transmute(Seral, Resistance, SourceWt,
            `Published (ha)` = num(ha_pub), `Corrected (ha)` = num(ha_cor),
            `Change (ha)` = num(d), `Change (%)` = num(pct)) |>
  styled(caption = paste(
    "Final resultant by seral stage. Resistance and source weight are assigned from the seral",
    "stage, so a stage that changes area moves that area between resistance classes."
  ), booktabs = TRUE)
```

```{r patches-missing, eval = !ok_patches, results = "asis"}
cat("_The corrected `forest_patches_final` was not available when this was rendered._\n")
```

## Patch statistics

Patch statistics now apply a 1 ha floor, matching the CEF Biodiversity Protocol threshold for
residual patches and the threshold arcpy `Eliminate` applies.
This is a reporting filter only: it never changes the layers the connectivity analysis is built
from, and the excluded count and area are reported alongside.

```{r patch-stats, results = "asis"}
pub_stats <- file.path(PUB_META, "Quesnel_TSA_seral_patch_stats.csv")
cor_stats <- file.path(proj, "Outputs", "Quesnel_TSA_seral_patch_stats.csv")
if (have(pub_stats, cor_stats)) {
  p <- read.csv(pub_stats); q <- read.csv(cor_stats)
  keep <- intersect(names(p), names(q))
  bind_rows(mutate(p[, keep], run = "published"), mutate(q[, keep], run = "corrected")) |>
    styled(caption = "Seral patch statistics, published vs corrected.", booktabs = TRUE) |>
    cat(sep = "\n")
} else {
  cat("_Patch statistics not available._")
}
```

# Figures

The two maps below are rendered by the same function from the same code path, so differences
between them are differences in the data and not in how they were drawn.

```{r map-before, eval = ok_patches, fig.cap = "Published patches (January 2026). Old is thin and fragmented; Mature is near-continuous.", out.width = "100%"}
targets::tar_source()
f <- file.path(FIGDIR, "patches_published.png")
if (!file.exists(f)) {
  plot_forest_disturbance_seral(b, tar_read_raw("Quesnel_TSA"),
                                "patches_published.png", outdir = FIGDIR)
}
include_graphics(f, rel_path = FALSE)
```

```{r map-after, eval = ok_patches, fig.cap = "Corrected patches (August 2026).", out.width = "100%"}
f <- file.path(FIGDIR, "patches_corrected.png")
if (!file.exists(f)) {
  plot_forest_disturbance_seral(a, tar_read_raw("Quesnel_TSA"),
                                "patches_corrected.png", outdir = FIGDIR)
}
include_graphics(f, rel_path = FALSE)
```

# Runtime and resources

## Runtime

Times are CPU seconds as recorded by `targets`, summed over branches, so they exceed wall clock
wherever branches ran in parallel.
They are grouped by workflow stage rather than by target: the rewrite split, merged and removed
targets, so individual names do not line up between the two runs, and several stages now do in one
target what previously took three.

Two things have to be filtered out or the comparison is meaningless.
The store keeps metadata for targets that were *not rebuilt*, which would compare the published run
against itself; and it also keeps metadata for targets the rewrite *removed*, which counts
eliminated work as though it still ran.
Totals below are filtered against `tar_manifest()` -- the targets that exist in the pipeline today
-- with removed targets attributed to the stage they used to serve, so each row shows the net effect
of the change rather than a deletion and an addition side by side.

```{r runtime-stage, results = "asis"}
## Group targets into workflow stages. Order matters: first match wins.
stage_of <- function(x) {
  dplyr::case_when(
    grepl("^omniscape", x) ~ "Omniscape configuration",
    grepl("interpatch_distances|quantiles_|forest_patches_chunks|patches_matold", x) ~
      "Interpatch distances",
    grepl("^(resistance|sourcewt)_", x) ~ "Resistance and source-weight surfaces",
    grepl("^patches_|^patch_summary|^forest_patches_final", x) ~ "Patch construction",
    grepl("^forest_disturbance_seral", x) ~ "Seral classification",
    grepl("^VRI_BECNDT", x) ~ "VRI x NDT-BEC overlay",
    grepl("^(readme|reproducibility_receipt)$", x) ~ "Reporting",
    TRUE ~ "Input layers and study area"
  )
}

roll <- function(store) {
  m <- tar_meta(store = store, fields = c("name", "seconds", "parent"))
  br <- m |> filter(!is.na(parent)) |> group_by(name = parent) |>
    summarise(seconds = sum(seconds, na.rm = TRUE), .groups = "drop")
  tp <- m |> filter(is.na(parent), !is.na(seconds)) |> transmute(name, seconds)
  bind_rows(br, tp) |> group_by(name) |> summarise(seconds = max(seconds), .groups = "drop")
}

if (dir.exists(PUB_META)) {
  live <- tar_manifest()$name
  o <- roll(PUB_META) |> rename(pub = seconds)
  n <- roll(file.path(proj, "_targets")) |> filter(name %in% live) |> rename(cor = seconds)

  by_stage <- full_join(o, n, by = "name") |>
    mutate(stage = stage_of(name)) |>
    group_by(stage) |>
    summarise(pub = sum(pub, na.rm = TRUE) / 3600,
              cor = sum(cor, na.rm = TRUE) / 3600, .groups = "drop") |>
    mutate(chg = cor - pub) |>
    arrange(chg)

  tot <- summarise(by_stage, stage = "**Total**", pub = sum(pub), cor = sum(cor), chg = sum(chg))

  bind_rows(by_stage, tot) |>
    transmute(
      Stage = stage,
      Published = sprintf("%.1f h", pub),
      Corrected = sprintf("%.1f h", cor),
      Change = ifelse(abs(chg) < 0.05, "--", sprintf("%+.1f h", chg))
    ) |>
    styled(caption = "Pipeline cost by workflow stage, published against corrected (CPU hours).",
          booktabs = TRUE, row.names = FALSE, align = "lrrr") |>
    cat(sep = "\n")
}
```

**A full build costs about 90 CPU hours less than the published one**, despite the corrections
adding work, and nearly all of that comes from two stages.

*Interpatch distances* (-50 h) lost `all_interpatch_distances_chunks`, which held the distances in
memory as a list of chunks; they are now written straight to a parquet dataset.
The quantile step over that dataset was itself rewritten after it was OOM-killed six hours into a
run: asking DuckDB for five quantiles as five aggregates made it keep five sorted states at once and
peak at 336 GB, where one `quantile_cont()` call over a list of probabilities peaks at 28.7 GB and
returns bit-identical values.

*Patch construction* (-40 h) lost `patches_union_final_3` (33.5 h to change 12 polygons out of
162,905, and zero area), and `patches_union_final` itself went from 5 h 52 m to 18 m by tiling.

The one stage that got *slower* is the VRI x NDT-BEC overlay (+0.5 h).
That is the price of correctness: the arcpy LENGTH sliver rule is more expensive than the area rule
it replaced, and it now runs on more and smaller polygons because the seral layer no longer
collapses duplicates.

These are CPU hours, not wall clock.
Wall clock is not comparable between the two runs -- the published one was driven by hand, and the
corrected one shared a machine with an unrelated job.

## Tiling the final resultant

`patches_union_final` is the arcpy step-6 overlay: the seral layer joined to both interior-forest
layers.
It ran as a single process and took **5 h 51 m 58 s**.

The overlay is embarrassingly parallel over space, so it now branches over `study_area_tiles` -- the
same 8x8 grid the seral layer itself was built on -- and reads its base layer through GDAL rather
than out of the `targets` store, so a worker materialises only the polygons inside its own tile.

```{r tiling-cost, results = "asis"}
data.frame(
  Quantity = c("Runtime", "Branches", "Slowest branch", "Features", "Planar area"),
  Serial = c("5 h 51 m 58 s", "1", "--", "189,666", "2,473,003.272 ha"),
  Tiled = c("17 m 48 s", "49, all succeeding", "2 m 48 s", "192,683", "2,473,003.272 ha"),
  check.names = FALSE
) |>
  styled(caption = "Building the final resultant, serial against tiled.", booktabs = TRUE,
         row.names = FALSE, label = "tiling-cost") |>
  cat(sep = "\n")
```

**19.8x faster, with the area preserved exactly.**

Reusing the *same* grid is load-bearing, not a convenience.
The seral layer already carries vertices lying bit-exactly on these seam coordinates, so re-cutting
on the same grid intersects at existing vertices and GEOS invents no new ones; the halves of a split
polygon abut exactly and dissolve back together.
A finer grid would cut at fresh coordinates and forfeit that.

Cropping splits polygons at the seams, so the resultant carries 3,017 more features (+1.59%).
That is a change to a shipped artefact -- `forest_patches_final.gpkg` -- even though no hectare
moves.
It does not propagate: `calc_matold()` dissolves before the interpatch distances are measured.

The verification that matters is the distance distribution, since that is what sets the Omniscape
radii:

```{r tiling-quantiles, results = "asis"}
data.frame(
  Quantile = c("25%", "50%", "75%", "90%", "99%", "100%"),
  Serial = c(0.372, 12.500, 39.787, 91.901, 330.123, 2566.803),
  Tiled = c(0.464, 12.603, 39.899, 92.131, 330.685, 2566.803)
) |>
  transform(Difference = sprintf("%+.3f", Tiled - Serial)) |>
  styled(caption = paste(
    "Nearest-neighbour distance quantiles in metres, serial against tiled.",
    "The 100% row is the one the Omniscape radius is derived from."
  ), booktabs = TRUE, row.names = FALSE, label = "tiling-quantiles") |>
  cat(sep = "\n")
```

Every quantile agrees to under a metre, and q100 -- the one `omniscape_config_nndist` actually
consumes -- is identical.
The radius is unchanged either way.

Dissolving does move slightly: `patches_matold` comes out 68,255 patches tiled against 68,401
serial.
This is the dissolve, not the tiling.
Before the dissolve the two layers agree to **0.000000 ha**; after it, the *serial* dissolve has
gained 0.798 ha, which a dissolve cannot do.
`terra::aggregate()` is simply not exact on a layer of this size -- the same layers self-overlap by
about 10 ha out of 2.47 Mha in both builds.

## The interior-forest overlay, and what tiling it would save

The final resultant was not the only overlay of this shape.
`patches_interior_forest_old` and `patches_interior_forest_mature_old` erase a buffer mask from the
old and mature+old patch layers, and between them they cost **8.1 CPU hours** in the corrected run --
3 h 07 m and 5 h 00 m respectively.
They were the largest remaining single-target costs once the resultant was tiled.

The cause is structural rather than incidental.
`patches_create_erase_mask()` rbinds the four buffer layers and dissolves them with
`terra::aggregate()`, which leaves the mask as a **single multipolygon covering 2,401,307.9 ha** --
the whole study area.
The cost of a difference is superlinear in the complexity of the layer being subtracted, so each of
the 42,764 patches was differenced against study-area-wide geometry no matter how small the patch
itself was.

Cropping both sides to a tile first shrinks the mask each difference sees to tile size.
Measured on Quesnel's own inputs, reusing the same 49-tile grid as the resultant and for the same
reason given above.
This has **not** been applied to the pipeline in this branch; the figures below are from a prototype
run against the stored inputs, kept here because they establish the size of the remaining cost and
the shape of the fix.

```{r interior-tiling-cost, results = "asis"}
data.frame(
  Quantity = c("Runtime", "Tiles", "Slowest tile", "Features", "Planar area"),
  `As built` = c("3 h 07 m 34 s", "1", "--", "10,263", "167,496.729062 ha"),
  Prototype = c("6 m 08 s", "49, all succeeding", "1 m 27 s", "10,263", "167,496.729059 ha"),
  check.names = FALSE
) |>
  styled(caption = paste(
    "Building the old interior-forest layer as built, against a tiled prototype.",
    "Sequential in both cases, so the tiled column is compute, not parallelism."
  ), booktabs = TRUE, row.names = FALSE, label = "interior-tiling-cost") |>
  cat(sep = "\n")
```

**30.5x faster, with the same 10,263 features and the area agreeing to 1.5e-11 relative** -- three
parts in a hundred billion, which is floating-point noise rather than a difference.
Unlike the resultant, the feature count does not change: the layer is dissolved after the tiles are
combined rather than per tile, so halves split at a seam are rejoined before anything downstream
sees them.

Tiling is only valid because this overlay is *local*: which part of a patch survives depends solely
on mask geometry near that patch.
The same treatment would be wrong for the dissolve or the interpatch distances, where a feature's
result depends on the whole layer.

The per-tile spread is the useful diagnostic: 0.2 s at the fastest, 2.8 s median, 87.5 s at the
slowest.
A factor of 31 between median and worst says cost is driven by local mask complexity rather than by
patch count, which is what makes the untiled form scale so badly -- every patch pays the worst case.

### Why this matters beyond Quesnel

Quesnel is the smallest of the Cariboo districts this workflow is intended to cover.
Cariboo-Chilcotin is roughly **2.4 times** its area, and the costs above do not scale with area:
the mask for a larger district is both larger and more intricate, and the superlinearity is in mask
complexity, not in patch count.
At the untiled cost profile, the interior-forest pair alone would plausibly dominate an entire
district build, and it is not the only overlay of this shape in the workflow.

This retooling is therefore a **prerequisite** for extending the analysis beyond Quesnel rather
than an optimisation of it.
The arithmetic is unforgiving in the other direction too: a step that takes a day is not merely slow,
it is a step that cannot be retried casually when something downstream needs changing, and this
report exists because several things did need changing.

To be unambiguous about what is and is not in this branch: every number reported elsewhere in this
document was produced by the pipeline as it stands, with this overlay untiled.
The prototype above reproduces the stored output exactly -- same 10,263 features, area agreeing to
1.5e-11 -- but it is a measurement, not a build.

The mechanism is available rather than hypothetical: `spatialutils` 0.0.1, which this branch pins,
provides `erase_polygons_tiled()` along with the guards `erase_polygons()` already carried.
What remains is to route this overlay through it, which is a change to
`patches_create_interior_forest()` and therefore invalidates every stored target downstream of it --
a deliberate decision to take on its own rather than a side effect of a performance change.

## Memory

Measured by sampling the resident set size of every project R process every 30 seconds across the
rerun, rather than estimated.

```{r memory}
data.frame(
  Measure = c("Peak aggregate RSS", "Largest single process", "Typical during tiled overlays",
              "terra ceiling per worker (`BC_CONN_MEMFRAC=0.01`)"),
  Value = c("110.0 GB", "41.1 GB", "~1.5 GB per branch", "~10 GB"),
  stringsAsFactors = FALSE
) |>
  styled(caption = paste(
    "Memory use of the data-preparation pipeline with 8 workers.",
    "Omniscape is excluded: it is opt-in and needs hundreds of GB."
  ), booktabs = TRUE)
```

# Implications for Omniscape

Omniscape consumes exactly two rasters: the composite resistance surface and the composite source
weights.
Both are assigned from `Seral`, so any area that changes seral stage moves between resistance
classes, and the connectivity result changes with it.

This section compares what Omniscape read and what it produced, before and after the corrections.
Both sides are still on disk: the published inputs were copied aside before the rerun overwrote
them, and the published runs survived because the corrections changed the run names (see
"Run naming").

## What changed in the inputs

```{r omniscape-rasters}
RAST_PUB <- file.path(PUBLISHED, "rasters")
RAST_CUR <- file.path(proj, "Data", "processed", "rasters")

## The corrected composite is written by the `resistance_composite` target; until that has been
## rebuilt the file on disk is still the published one, and comparing it with its own backup would
## report no change at all.
ok_rast <- dir.exists(RAST_PUB) && is_rebuilt("resistance_composite")

## Read each raster once into a plain vector and do the arithmetic in R. Repeated
## `terra::global()` calls re-read the file for every statistic, which at 30 m -- 69.2 M cells --
## turned this comparison from seconds into minutes.
##
## A cell counts as changed if its value differs OR its NA status differs. The second case is not
## a detail here: the forest layers are NA off-patch, so a cell that leaves the patch set goes
## from a number to NA, and a value-only comparison would miss it entirely.
cmp_rast <- function(name, res) {
  f <- sprintf("%s_%d.tif", name, res)
  ra <- terra::rast(file.path(RAST_PUB, f))
  rb <- terra::rast(file.path(RAST_CUR, f))
  if (!terra::compareGeom(ra, rb, stopOnError = FALSE)) return(NULL)
  va <- terra::values(ra, mat = FALSE)
  vb <- terra::values(rb, mat = FALSE)
  na_a <- is.na(va); na_b <- is.na(vb)
  both <- !na_a & !na_b
  cellha <- prod(terra::res(ra)) / 1e4
  out <- data.frame(
    layer = name, res = res,
    ha_chg = (sum(na_a != na_b) + sum(both & va != vb)) * cellha,
    ha_lost = sum(!na_a & na_b) * cellha,
    ha_gained = sum(na_a & !na_b) * cellha,
    mean_pub = mean(va[both]), mean_cor = mean(vb[both]),
    stringsAsFactors = FALSE
  )
  rm(va, vb, na_a, na_b, both); gc(verbose = FALSE)
  out
}

LAYERS <- c("resistance_forest", "resistance_composite",
            "sourcewt_forest", "sourcewt_composite")
LAYER_LABELS <- c(resistance_forest = "Resistance, forest",
                  resistance_composite = "Resistance, composite",
                  sourcewt_forest = "Source weight, forest",
                  sourcewt_composite = "Source weight, composite")

## Prose below quotes these inline. Without NA defaults a missing artefact would turn a chunk that
## is merely skipped into a render that fails, several pages from the chunk that did not run.
tot_ha <- NA_real_
ca <- cb <- stats::setNames(rep(NA_real_, 6), c("1", "250", "500", "750", "1000", "mixed"))
pair_pub <- pair_cor <- NA_real_
```

```{r omniscape-input-diff, eval = ok_rast, results = "asis"}
inp <- do.call(rbind, lapply(c(30, 90), function(r) {
  do.call(rbind, lapply(LAYERS, cmp_rast, res = r))
}))

## Study-area total, for the share-changed column. Taken from the composite, which unlike the
## forest layers is complete over the study area -- it has no NA cells at all.
tot_ha <- {
  r <- terra::rast(file.path(RAST_CUR, "resistance_composite_30.tif"))
  sum(!is.na(terra::values(r, mat = FALSE))) * prod(terra::res(r)) / 1e4
}

dp <- ifelse(grepl("sourcewt", inp$layer), 4, 2)
data.frame(
  Layer = unname(LAYER_LABELS[inp$layer]),
  `Res (m)` = inp$res,
  `Area changed (ha)` = num(inp$ha_chg, 0),
  `Share of study area` = paste0(num(100 * inp$ha_chg / tot_ha, 2), " %"),
  `Mean, published` = mapply(num, inp$mean_pub, dp),
  `Mean, corrected` = mapply(num, inp$mean_cor, dp),
  check.names = FALSE
) |>
  styled(caption = paste(
    "Change in the rasters Omniscape reads. The forest layers are NA off-patch; the composites",
    "fill those cells with the matrix value and are complete over the study area.",
    "Means are taken over cells present in both versions."
  ), booktabs = TRUE, row.names = FALSE, label = "omniscape-inputs") |>
  cat(sep = "\n")
```

```{r omniscape-pending, eval = !ok_rast, results = "asis"}
cat("_The corrected resistance surface had not been rebuilt when this was rendered._\n")
```

The 30 m and 90 m rows are independent measurements of the same change on the ground, and they
agree to better than 0.1% on every layer.
That is the check that this is a real difference in the patch set and not an artefact of one grid.

Two things in that table need separating.

**The composite means move in the direction of less connectivity**: mean resistance rises about
1.4% and mean source weight falls about 4.8%.
The corrected landscape is slightly harder to move through than the published one said it was.

**The forest layers change over more ground than the composites do**, which is not a contradiction.
The forest layer is NA wherever there is no patch, so it registers a change both when a cell's
resistance class moves and when the cell enters or leaves the patch set; in the composite the
off-patch cells carry the matrix value instead, so only the first kind appears there as a change.

```{r omniscape-extent, eval = ok_rast, results = "asis"}
fe <- subset(inp, layer == "resistance_forest" & res == 30)
data.frame(
  Measure = c("In the published patch layer, not the corrected one",
              "In the corrected patch layer, not the published one",
              "Net change in patch extent"),
  Value = c(paste(num(fe$ha_lost, 0), "ha"),
            paste(num(fe$ha_gained, 0), "ha"),
            paste(num(fe$ha_gained - fe$ha_lost, 0), "ha")),
  stringsAsFactors = FALSE
) |>
  styled(caption = "Change in the extent of the forest patch layer, at 30 m.",
         booktabs = TRUE, label = "omniscape-extent") |>
  cat(sep = "\n")
```

The loss is almost entirely one-directional, and it is the raster expression of the patch-area
change reported under "Forest patches".

### Where the land went

Resistance is assigned in discrete classes, so the composite can be compared class by class.

```{r omniscape-classes, eval = ok_rast, results = "asis"}
cellha <- 30 * 30 / 1e4

## The nominal classes are what `Resistance` is assigned from. Everything else in the raster is a
## cell that `terra::rasterize(fun = mean)` averaged because more than one polygon covered it,
## and those are pooled into one row -- they are the subject of the paragraph below.
##
## `terra::freq()` counts values in C++ without materialising 69.2 M doubles in R, which matters
## here because the transition table below needs both vectors in memory at once anyway.
##
## `digits` defaults to 0, which ROUNDS the values before counting -- it silently merges 125.5 into
## 126, and the mixture values this section is about stop existing. The nominal classes are whole
## numbers so the totals come out right either way, which is what makes the default dangerous here.
NOMINAL <- c(1, 250, 500, 750, 1000)
FREQ_DIGITS <- 4
cls <- function(f) {
  tb <- terra::freq(terra::rast(f), digits = FREQ_DIGITS)
  k <- ifelse(tb$value %in% NOMINAL, as.character(tb$value), "mixed")
  out <- tapply(tb$count * cellha, factor(k, levels = c(as.character(NOMINAL), "mixed")), sum)
  out[is.na(out)] <- 0
  out
}
ca <- cls(file.path(RAST_PUB, "resistance_composite_30.tif"))
cb <- cls(file.path(RAST_CUR, "resistance_composite_30.tif"))

## The two mixture values that are exactly the average of an adjacent pair of seral classes:
## 125.5 = (1 + 250) / 2 and 375 = (250 + 500) / 2. They are the ones the paragraph below is about.
pair_ha <- function(f) {
  tb <- terra::freq(terra::rast(f), digits = FREQ_DIGITS)
  sum(tb$count[abs(tb$value - 125.5) < 1e-6 | abs(tb$value - 375) < 1e-6]) * cellha
}
pair_pub <- pair_ha(file.path(RAST_PUB, "resistance_composite_30.tif"))
pair_cor <- pair_ha(file.path(RAST_CUR, "resistance_composite_30.tif"))

data.frame(
  `Resistance` = c(paste0(NOMINAL, c(" (best)", "", "", "", " (matrix)")), "mixed cells"),
  `Source weight` = c("1.00", "0.75", "0.50", "0.25", "0.00", "--"),
  `Published (ha)` = num(as.numeric(ca), 0),
  `Corrected (ha)` = num(as.numeric(cb), 0),
  `Change (ha)` = paste0(ifelse(cb - ca >= 0, "+", ""), num(as.numeric(cb - ca), 0)),
  check.names = FALSE
) |>
  styled(caption = paste(
    "Area in each resistance class of the composite at 30 m, with the source weight assigned",
    "from the same seral class. Mixed cells are those covered by more than one polygon, which",
    "rasterize averages."
  ), booktabs = TRUE, row.names = FALSE, label = "omniscape-classes") |>
  cat(sep = "\n")
```

The correction **polarises** the landscape.
The best class more than doubles and the matrix grows, both at the expense of the middle: the
250 class alone loses about a quarter of a million hectares, split between the two extremes.
This is why the composite means barely move while so much ground changes class -- the gains at
each end largely cancel in the mean.

It also explains the last row, which is worth dwelling on because it is independent evidence.

`create_resistance_raster()` rasterises with `fun = mean`, and `terra::rasterize()` assigns a cell
from every polygon covering its centre.
A cell written as 125.5 is therefore one whose centre fell inside a resistance-1 polygon **and** a
resistance-250 polygon at the same time -- 125.5 being the average of the two.
Cells like that can only exist where the patch layer overlaps itself, which is precisely what the
`terra::union()` defect produced.
The composite is assembled with `pmax()` and does no averaging of its own, so every mixed cell in
it was already mixed in one of the layers it reads.

Mixed cells fall from `r num(ca[["mixed"]], 0)` ha to `r num(cb[["mixed"]], 0)` ha.
The two values that are exactly the average of an adjacent pair of seral classes -- 125.5 and 375 --
account for `r num(pair_pub, 0)` ha of the published total and `r num(pair_cor, 0)` ha of the
corrected one.
That is raster-level confirmation, arrived at without going near the polygons, that the overlapping
geometry is gone.
The `r num(cb[["mixed"]], 0)` ha that remain are `r num(100 * cb[["mixed"]] / tot_ha, 2)`% of the
study area; overlap has not been eliminated outright, but what is left is small enough not to move
any figure in this report.

```{r omniscape-transitions, eval = ok_rast, results = "asis"}
## Tabulate only the cells that actually changed. A full cross-tabulation of both vectors would
## factor 69.2 M values against ~300 levels twice; restricted to the changed cells it is 18x
## smaller and the result is identical, because the diagonal is dropped anyway.
va <- terra::values(terra::rast(file.path(RAST_PUB, "resistance_composite_30.tif")), mat = FALSE)
vb <- terra::values(terra::rast(file.path(RAST_CUR, "resistance_composite_30.tif")), mat = FALSE)
ch <- which(va != vb)
tr <- as.data.frame(table(from = va[ch], to = vb[ch]), stringsAsFactors = FALSE)
tr <- tr[tr$Freq > 0, ]
tr$ha <- tr$Freq * cellha
tr <- head(tr[order(-tr$ha), ], 6)

## These are class labels, so they get no thousands separator and no forced decimal: "1000",
## not "1,000.0", and "125.5" kept as it is.
lbl <- function(x) ifelse(x == round(x), sprintf("%.0f", x), sprintf("%.1f", x))

data.frame(
  `Published resistance` = lbl(as.numeric(tr$from)),
  `Corrected resistance` = lbl(as.numeric(tr$to)),
  `Area (ha)` = num(tr$ha, 0),
  check.names = FALSE
) |>
  styled(caption = "The six largest resistance-class transitions in the composite at 30 m.",
         booktabs = TRUE, row.names = FALSE, label = "omniscape-transitions") |>
  cat(sep = "\n")

## `results = "asis"` prints whatever the chunk returns, and gc() returns its summary matrix.
rm(va, vb, ch); invisible(gc(verbose = FALSE))
```

The two largest movements are in opposite directions and comparable in size.
Land the published run scored as ordinary forest is best-quality habitat in the corrected one, and
a nearly equal area the published run scored as forest patch is not in a patch at all.
The defect was not a uniform over- or under-count that a single correction factor could absorb; it
misplaced habitat in both directions.

## Run naming

Omniscape run directories are stamped with `OMNISCAPE_VINTAGE`, set in `_targets.R`.
It was previously a hardcoded `"2026-01-23"`, which is the vintage of the published runs.
Because `run_omniscape()` deletes an existing output directory before running, a corrected run whose
radius matched a published one would have destroyed it -- and there are 17 GB of published results,
four of them under that prefix, including the run the report was written from.
Corrected runs are stamped `2026-08-26`.

## Radii

The Omniscape `radius`, `block_size` and `buffer` are all in **pixels**, not map units.
Radii are derived from quantiles of the interpatch distance distribution, so the corrected patches
imply new radii and hence new run names.

**Both radii move**, and by more than a pixel.

The published values are recoverable independently of any stored quantile, because
`write_omniscape_config()` stamps the radius into the run directory name: the `2026-01-23` vintage
wrote `p30_r1507_bs151` (regional) and `p30_r58_bs7` / `p90_r20_bs3` (local).

```{r radii, results = "asis"}
ok_q <- is_rebuilt("quantiles_nn_dists") && is_rebuilt("quantiles_all_dists")
qn <- try(tar_read_raw("quantiles_nn_dists"), silent = TRUE)
qa <- try(tar_read_raw("quantiles_all_dists"), silent = TRUE)

if (ok_q && !inherits(qn, "try-error") && !inherits(qa, "try-error")) {
  ## `use_radius()` takes the quantile in METRES and returns pixels, so the pin is resolution-
  ## independent: the same distance gives a different pixel count at 30 m and at 90 m.
  px <- function(m, res) ceiling(m / res)

  nn100 <- as.numeric(qn[["100%"]])
  all25 <- as.numeric(qa[["25%"]])

  ## published metres, back-derived from the run names above (radius px * pixel size)
  pub <- data.frame(
    Quantity = c("Regional (all-pairs q25)", "Local (nearest-neighbour q100)"),
    `Published m` = c(45202, 1726),
    `Corrected m` = c(all25, nn100),
    `px at 30 m` = c(sprintf("1507 -> %d", px(all25, 30)), sprintf("58 -> %d", px(nn100, 30))),
    `px at 90 m` = c(sprintf("503 -> %d", px(all25, 90)), sprintf("20 -> %d", px(nn100, 90))),
    check.names = FALSE
  )
  pub$`Published m` <- num(pub$`Published m`, 0)
  pub$`Corrected m` <- num(pub$`Corrected m`, 0)

  cat(styled(pub, caption = "Radii before and after the corrections.", booktabs = TRUE,
             row.names = FALSE, label = "radii-change"), sep = "\n")
  cat("\n\n")

  show_q <- function(x, cap, lbl) {
    d <- as.data.frame(x)
    if (nrow(d) > 20) d <- d[seq(1, nrow(d), length.out = 20), , drop = FALSE]
    ## an explicit label per table: three tables come out of this one chunk, and knitr labels
    ## them all with the chunk name otherwise, so every cross-reference resolves to the last one
    cat(styled(d, caption = cap, booktabs = TRUE, row.names = FALSE, label = lbl), sep = "\n")
    cat("\n\n")
  }
  show_q(qn, "Nearest-neighbour interpatch distance quantiles (corrected).", "quantiles-nn")
  show_q(qa, "All-pairs interpatch distance quantiles (corrected).", "quantiles-all")
} else {
  cat("_Distance quantiles had not been rebuilt when this was rendered;",
      "the radii for the corrected runs are therefore not yet known._\n")
}
```

The regional radius falls by 5.2% and the local one rises by 49%.
Neither is within the roughly one-pixel slack that would leave the pixel count unchanged, so every
corrected run needs its own configuration -- which is what the four `2026-08-26` directories are.

The direction differs because the two quantiles measure different things.
The all-pairs q25 fell because the corrected patch set covers less area overall; the
nearest-neighbour q100 is a **maximum**, and maxima are sensitive to the patch-construction defect
in a way that interior quantiles are not.

This is entirely a consequence of the corrections, not of tiling: the serial and tiled builds give
q100 identical to the millimetre (see "Tiling the final resultant").

## What changed in the outputs

```{r omniscape-outputs-setup}
## Published run -> corrected run, matched by scale and resolution. The radii differ because
## they are derived from the corrected interpatch distances; see "Radii".
RUN_PAIRS <- list(
  list(scale = "Regional", res = 30, pub = "2026-01-23_p30_r1507_bs151", cor = "2026-08-26_p30_r1429_bs143"),
  list(scale = "Regional", res = 90, pub = "2026-01-23_p90_r503_bs51",   cor = "2026-08-26_p90_r477_bs49"),
  list(scale = "Local",    res = 30, pub = "2026-01-23_p30_r58_bs7",     cor = "2026-08-26_p30_r86_bs9"),
  list(scale = "Local",    res = 90, pub = "2026-01-23_p90_r20_bs3",     cor = "2026-08-26_p90_r29_bs3")
)
OUT_LAYERS <- c("cum_currmap", "flow_potential", "normalized_cum_currmap")

run_tif <- function(run, layer) file.path(proj, "Outputs", run, paste0(layer, ".tif"))

## As for the inputs: prose below quotes these inline, so they need to exist even when the
## comparison is skipped.
nz_stat <- stats::setNames(rep(NA_real_, 11), c(
  "mean_lo", "mean_hi", "r_lo", "r_hi", "w5_lo", "w5_hi",
  "ha_lo", "ha_hi", "pct_lo", "pct_hi", "chan_max"))
raw_gain <- NA_real_

## Radius is stamped into the run directory name, so the change is recoverable from the names
## themselves rather than having to be restated in the prose and kept in step by hand.
radius_of <- function(run) as.numeric(sub(".*_r([0-9]+)_.*", "\\1", run))
local_radius_gain <- min(vapply(
  Filter(function(p) p$scale == "Local", RUN_PAIRS),
  function(p) 100 * (radius_of(p$cor) / radius_of(p$pub) - 1), numeric(1)
))

ok_runs <- all(vapply(RUN_PAIRS, function(p) {
  all(file.exists(vapply(OUT_LAYERS, function(l) c(run_tif(p$pub, l), run_tif(p$cor, l)), character(2))))
}, logical(1)))

## As for the inputs: one read into a plain vector per raster, then arithmetic in R. Twelve pairs
## at up to 69.2 M cells is 6.8 GB of I/O however it is done, but this way it is done once.
cmp_run <- function(p, layer) {
  ra <- terra::rast(run_tif(p$pub, layer))
  rb <- terra::rast(run_tif(p$cor, layer))
  if (!terra::compareGeom(ra, rb, stopOnError = FALSE)) return(NULL)
  a <- terra::values(ra, mat = FALSE); b <- terra::values(rb, mat = FALSE)
  ok <- !is.na(a) & !is.na(b)
  a <- a[ok]; b <- b[ok]
  d <- b - a
  cellha <- prod(terra::res(ra)) / 1e4
  out <- data.frame(
    scale = p$scale, res = p$res, layer = layer,
    mean_pub = mean(a), mean_cor = mean(b),
    r = suppressWarnings(stats::cor(a, b)),
    ## share of cells whose value moved by no more than 5% -- a cell-level measure of agreement
    ## that, unlike the correlation, is not dominated by the long right tail of current values
    within5 = 100 * mean(abs(d) <= 0.05 * pmax(abs(a), 1e-12)),
    stringsAsFactors = FALSE
  )
  if (layer == "normalized_cum_currmap") {
    ## 1.0 is the only non-arbitrary breakpoint in a normalized current map: it is exactly the
    ## point where current equals flow potential. Above it a cell is channelized, below impeded,
    ## so a cell that crosses it changes its interpretation and not just its value.
    out$ha_cross <- sum((a > 1) != (b > 1)) * cellha
    out$pct_cross <- 100 * mean((a > 1) != (b > 1))
    out$chan_pub <- 100 * mean(a > 1)
    out$chan_cor <- 100 * mean(b > 1)
  } else {
    out$ha_cross <- NA; out$pct_cross <- NA; out$chan_pub <- NA; out$chan_cor <- NA
  }
  rm(a, b, d); gc(verbose = FALSE)
  out
}
```

The published and corrected runs differ in **two** ways at once, and the artefacts on disk cannot
separate them: the inputs changed, and the radii derived from those inputs changed with them.
Both are consequences of the same fix, so the totals below are the right answer to "what does the
correction do to the published maps".
They are *not* an attribution of that difference to the input change alone; isolating it would take
a further run of the corrected inputs at the published radius, which has not been made.

This matters most for the raw current maps.
`cum_currmap` and `flow_potential` both scale with the size of the moving window, and the local
radius rose by at least `r num(local_radius_gain, 0)`%, so much of any change in those two layers
is the window rather than the landscape.
`normalized_cum_currmap` divides one by the other, which removes most of that effect, and it is
also the layer the project's own summaries are built on.
**It is the layer to read the comparison from.**


```{r omniscape-outputs, eval = ok_runs, results = "asis"}
outs <- do.call(rbind, unlist(
  lapply(RUN_PAIRS, function(p) lapply(OUT_LAYERS, cmp_run, p = p)),
  recursive = FALSE
))

pretty_layer <- c(cum_currmap = "Cumulative current",
                  flow_potential = "Flow potential",
                  normalized_cum_currmap = "Normalized current")

data.frame(
  Run = paste0(outs$scale, ", ", outs$res, " m"),
  Layer = unname(pretty_layer[outs$layer]),
  `Mean, published` = num(outs$mean_pub, 3),
  `Mean, corrected` = num(outs$mean_cor, 3),
  `Change` = paste0(ifelse(outs$mean_cor >= outs$mean_pub, "+", ""),
                    num(100 * (outs$mean_cor - outs$mean_pub) / outs$mean_pub, 1), " %"),
  `Correlation` = num(outs$r, 3),
  `Cells within 5%` = paste0(num(outs$within5, 1), " %"),
  check.names = FALSE
) |>
  styled(caption = paste(
    "Omniscape output before and after the corrections, over cells present in both runs.",
    "The raw current and flow potential also carry the change in radius; the normalized",
    "current largely does not."
  ), booktabs = TRUE, row.names = FALSE, label = "omniscape-outputs") |>
  cat(sep = "\n")

## Quoted inline in the prose that follows, so that the text cannot drift away from the table.
nzr <- subset(outs, layer == "normalized_cum_currmap")
dpct <- 100 * (nzr$mean_cor - nzr$mean_pub) / nzr$mean_pub
nz_stat[] <- c(min(dpct), max(dpct), min(nzr$r), max(nzr$r),
               min(nzr$within5), max(nzr$within5),
               min(nzr$ha_cross), max(nzr$ha_cross),
               min(nzr$pct_cross), max(nzr$pct_cross),
               max(abs(nzr$chan_cor - nzr$chan_pub)))
raw_gain <- min(100 * (subset(outs, scale == "Local" & layer != "normalized_cum_currmap")$mean_cor /
                       subset(outs, scale == "Local" & layer != "normalized_cum_currmap")$mean_pub - 1))
```

```{r omniscape-outputs-pending, eval = !ok_runs, results = "asis"}
cat("_Not every published and corrected run pair was on disk when this was rendered._\n")
```

The local runs bear that out: raw current and flow potential both rise by at least
`r num(raw_gain, 0)`% there, tracking the radius rather than anything about the landscape.
Read down the normalized-current rows instead.

The mean of the normalized current map moves by between `r num(nz_stat[["mean_lo"]], 1)`% and
`r num(nz_stat[["mean_hi"]], 1)`%.
Taken alone that reads as a negligible correction.
It is not, and the next two columns say why: the cell-by-cell correlation between the published and
corrected maps is `r num(nz_stat[["r_lo"]], 2)` to `r num(nz_stat[["r_hi"]], 2)`, and only
`r num(nz_stat[["w5_lo"]], 0)`% to `r num(nz_stat[["w5_hi"]], 0)`% of cells are within 5% of their
published value.
**The summary statistic is nearly unchanged while the map underneath it is substantially
rearranged** -- which is the same failure mode as the defects themselves, a plausible-looking number
standing in front of a wrong one.

```{r omniscape-crossings, eval = ok_runs, results = "asis"}
nz <- subset(outs, layer == "normalized_cum_currmap")
data.frame(
  Run = paste0(nz$scale, ", ", nz$res, " m"),
  `Channelized, published` = paste0(num(nz$chan_pub, 1), " %"),
  `Channelized, corrected` = paste0(num(nz$chan_cor, 1), " %"),
  `Area crossing (ha)` = num(nz$ha_cross, 0),
  `Share of study area` = paste0(num(nz$pct_cross, 1), " %"),
  check.names = FALSE
) |>
  styled(caption = paste(
    "Cells crossing the normalized-current value of 1, where current equals flow potential.",
    "A cell above 1 is channelized, below it impeded, so crossing changes the interpretation",
    "of that location and not only its value."
  ), booktabs = TRUE, row.names = FALSE, label = "omniscape-crossings") |>
  cat(sep = "\n")
```

Between `r num(nz_stat[["ha_lo"]], 0)` and `r num(nz_stat[["ha_hi"]], 0)` ha --
`r num(nz_stat[["pct_lo"]], 1)`% to `r num(nz_stat[["pct_hi"]], 1)`% of the study area -- changes
side of that line.
The *share* of the district that is channelized barely moves, by at most
`r num(nz_stat[["chan_max"]], 1)` percentage points.
So this is a redistribution rather than a level shift: the corrected run finds about as much
channelized land as the published one did, in different places.

For a report whose conclusions are about where the pinch points are, that is the material result.

### Zonal summaries

`zonal_summaries()` reduces the normalized current map to a mean per biodiversity feature class and
per Landscape Unit x BEC zone combination.
Those means, not the rasters, are what the reporting was written from, so they are the level at
which the correction actually reaches a conclusion.

```{r omniscape-zonal-setup}
zonal_of <- function(run) {
  f <- file.path(proj, "Outputs", run, "zonal_by_polys.csv")
  if (!file.exists(f)) return(NULL)
  z <- utils::read.csv(f, stringsAsFactors = FALSE)
  stats::setNames(z$mean_norm_cum_curr, z$zone)
}

zz <- do.call(rbind, lapply(RUN_PAIRS, function(p) {
  a <- zonal_of(p$pub); b <- zonal_of(p$cor)
  if (is.null(a) || is.null(b)) return(NULL)
  k <- intersect(names(a), names(b))
  data.frame(run = paste0(p$scale, ", ", p$res, " m"), zone = k,
             pub = unname(a[k]), cor = unname(b[k]), stringsAsFactors = FALSE)
}))

ok_zonal <- !is.null(zz) && nrow(zz) > 0
if (ok_zonal) {
  ## A handful of high-alpine LU x BEC combinations carry no current at all at the local radius,
  ## in both runs. 0 -> 0 is not a 0% change, it is no basis for a percentage, so those zones are
  ## dropped and counted rather than allowed to propagate a NaN into a median.
  zz$dpct <- ifelse(zz$pub == 0, NA_real_, 100 * (zz$cor - zz$pub) / zz$pub)
}
n_dropped <- if (ok_zonal) sum(is.na(zz$dpct) & grepl("^LU_", zz$zone)) else NA
```

```{r omniscape-zonal-features, eval = ok_zonal, results = "asis"}
FEATURES <- c(parks = "Parks", ogma = "OGMA", wha = "Wildlife habitat areas",
              mdwr = "Mule deer winter range", hvmw = "High-value moose wetlands",
              wetlands = "Wetlands")
fz <- subset(zz, zone %in% names(FEATURES) & run == "Regional, 30 m")
fz <- fz[order(-fz$dpct), ]

data.frame(
  Feature = unname(FEATURES[fz$zone]),
  `Published` = num(fz$pub, 3),
  `Corrected` = num(fz$cor, 3),
  `Change` = paste0(ifelse(fz$dpct >= 0, "+", ""), num(fz$dpct, 1), " %"),
  check.names = FALSE
) |>
  styled(caption = paste(
    "Mean normalized current within each biodiversity feature class, regional run at 30 m.",
    "A value above 1 means current concentrates there relative to flow potential."
  ), booktabs = TRUE, row.names = FALSE, label = "omniscape-zonal-features") |>
  cat(sep = "\n")
```

The features do not move together.
Parks, wildlife habitat areas, OGMA and mule deer winter range all gain; the two wetland classes
lose.
The complements tell the same story from the other side: every `non-` zone rises by between 1.4%
and 3.6%, so the corrected run puts relatively more current outside the designated features than
the published run did -- except in parks, where the gain inside is large enough to run the other
way.

Parks is the one to note.
A `r num(abs(subset(zz, zone == "parks" & run == "Regional, 30 m")$dpct), 1)`% change in the mean
normalized current of the protected-areas network is not a rounding correction, and it is the kind
of number that gets quoted on its own.

```{r omniscape-zonal-lu, eval = ok_zonal, results = "asis"}
lz <- subset(zz, grepl("^LU_", zone) & !is.na(dpct))
agg <- do.call(rbind, lapply(split(lz, lz$run), function(g) {
  data.frame(
    Run = g$run[1],
    `Zones` = nrow(g),
    `Median change` = paste0(ifelse(stats::median(g$dpct) >= 0, "+", ""),
                             num(stats::median(g$dpct), 1), " %"),
    `Median absolute` = paste0(num(stats::median(abs(g$dpct)), 1), " %"),
    `90th pct absolute` = paste0(num(stats::quantile(abs(g$dpct), 0.9), 0), " %"),
    `Zones moving > 10%` = num(sum(abs(g$dpct) > 10), 0),
    check.names = FALSE
  )
}))
agg <- agg[match(vapply(RUN_PAIRS, function(p) paste0(p$scale, ", ", p$res, " m"), ""), agg$Run), ]

cap <- paste(
  "Change in mean normalized current across the Landscape Unit by BEC zone summaries.",
  "The median zone barely moves; the spread around it is large."
)
## Say so in the caption rather than silently showing a smaller denominator than the reader expects.
if (isTRUE(n_dropped > 0)) {
  cap <- paste(cap, sprintf(
    paste("%d zone-and-run combinations are excluded: a few high-alpine units carry no current at",
          "all at the local radius, in both runs, and 0 to 0 is no basis for a percentage."),
    n_dropped))
}

cat(styled(agg, caption = cap, booktabs = TRUE, row.names = FALSE,
           label = "omniscape-zonal-lu"), sep = "\n")
```

```{r omniscape-zonal-extremes, eval = ok_zonal, results = "asis"}
ez <- subset(zz, grepl("^LU_", zone) & run == "Regional, 30 m" & !is.na(dpct))
ez <- ez[order(-abs(ez$dpct)), ][1:6, ]

## Zone keys are built as LU_<unit name>_<BEC zone>, and unit names themselves contain spaces,
## so split on the LAST underscore. Left as-is they also reach LaTeX full of escaped underscores.
data.frame(
  `Landscape unit and BEC zone` = sub("^LU_(.*)_([^_]+)$", "\\1 (\\2)", ez$zone),
  `Published` = num(ez$pub, 2),
  `Corrected` = num(ez$cor, 2),
  `Change` = paste0(ifelse(ez$dpct >= 0, "+", ""), num(ez$dpct, 0), " %"),
  check.names = FALSE
) |>
  styled(caption = paste(
    "The six Landscape Unit by BEC zone summaries that move most, regional run at 30 m."
  ), booktabs = TRUE, row.names = FALSE, label = "omniscape-zonal-extremes") |>
  cat(sep = "\n")
```

These are not small denominators inflating a percentage: the published means involved are between
0.8 and 3.1, comfortably in range.
They are zones whose character reverses.
A landscape unit the published run reported as one of the strongest current concentrations in the
district comes back as one of the weakest, and another makes the opposite journey.

That is the concrete form of the redistribution the raster comparison measured, and it is the
reason the corrected runs cannot be treated as a refinement of the published ones.
The district-wide totals are close enough that they would not, on their own, have prompted a
rerun; the zone-level conclusions drawn from the published maps do not survive it.

# Reproducing this report

This report is deliberately **not** a pipeline target: it compares two runs, one of which no longer
exists as a live `targets` store.
It reads:

- `Data/processed/published-2026-01-27/` -- the published patches layer, composite rasters and map,
  copied aside before the rerun overwrote them;
- the `targets` metadata snapshot of the published run;
- the live `_targets` store for the corrected values.

Render with:

The rendered PDF is written to `Outputs/`, not next to the source, so it lands on the shared volume
alongside the other results rather than inside the clone:

```{r render-cmd, echo = TRUE, eval = FALSE}
rmarkdown::render(
  "reports/corrections-2026-08.Rmd",
  output_file = "corrections-2026-08.pdf",
  output_dir = "Outputs"
)
```

```{r session}
data.frame(
  Item = c("R", "targets", "terra", "sf", "git branch", "git commit"),
  Value = c(
    as.character(getRversion()),
    as.character(packageVersion("targets")),
    as.character(packageVersion("terra")),
    as.character(packageVersion("sf")),
    system("git rev-parse --abbrev-ref HEAD", intern = TRUE),
    system("git rev-parse --short HEAD", intern = TRUE)
  ),
  stringsAsFactors = FALSE
) |>
  styled(caption = "Environment.", booktabs = TRUE)
```
