---
title: "Quesnel NRD connectivity: 30 m vs 90 m raster comparisons"
author: "Alex Chubaty"
date: last-modified
date-format: "[Last updated:] YYYY-MM-DD"
bibliography: ../citations/references.bib
csl: ../citations/ecology-letters.csl
format:
  pdf:
    output-file: resolution-equivalence.pdf
    documentclass: article
    geometry: margin=2.5cm
    toc: true
    toc-depth: 2
    number-sections: true
    fig-pos: "H"
    pdf-engine: pdflatex
    urlcolor: blue
    include-in-header:
      text: |
        \usepackage{float}
        \floatplacement{figure}{H}
        \floatplacement{table}{H}
        \usepackage{makecell}
        \usepackage{needspace}
        \let\oldsection\section
        \renewcommand{\section}{\needspace{5\baselineskip}\oldsection}
        \let\oldsubsection\subsection
        \renewcommand{\subsection}{\needspace{5\baselineskip}\oldsubsection}
        \clubpenalty=10000
        \widowpenalty=10000
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.pos = "H",
  out.extra = ""
)

find_project_root <- function(start = getwd()) {
  p <- normalizePath(start, mustWork = FALSE)
  while (!file.exists(file.path(p, "_targets.R")) && dirname(p) != p) {
    p <- dirname(p)
  }
  if (!file.exists(file.path(p, "_targets.R"))) {
    stop("could not locate the project root")
  }
  p
}

proj <- find_project_root()
knitr::opts_knit$set(root.dir = proj)

options(knitr.table.format = "latex")

library(dplyr)
library(knitr)

## Quesnel's Omniscape outputs moved under `Outputs/quesnel/` when the pipeline became
## multi-district. A clone from before that move still has them flat. Resolve either, so this
## document does not care which layout it is rendering against.
run_dir <- function(run) {
  cand <- c(
    file.path(proj, "Outputs", "quesnel", run),
    file.path(proj, "Outputs", run)
  )
  hit <- cand[dir.exists(cand)]
  if (!length(hit)) {
    return(NA_character_)
  }
  hit[[1]]
}

## The district boundary, for clipping. Both layouts again: `quesnel/` after the move, flat before.
##
## The UNBUFFERED boundary, deliberately. `Quesnel_TSA_studyarea_buffered.gpkg` sits beside it and
## is 3,580,800 ha against 2,083,306 ha; that buffer exists so Omniscape's moving window has room
## to work and is not part of the reporting area. Everything counted or mapped in this document is
## the district itself.
study_area_path <- function() {
  cand <- c(
    file.path(proj, "Data", "processed", "quesnel", "Quesnel_TSA_studyarea.gpkg"),
    file.path(proj, "Data", "processed", "Quesnel_TSA_studyarea.gpkg")
  )
  hit <- cand[file.exists(cand)]
  if (!length(hit)) NA_character_ else hit[[1]]
}

rmd_dir <- tryCatch(dirname(knitr::current_input(dir = TRUE)), error = function(e) NULL)
if (is.null(rmd_dir) || !nzchar(rmd_dir) || is.na(rmd_dir)) {
  rmd_dir <- file.path(proj, "reports")
}
FIGDIR <- normalizePath(file.path(rmd_dir, "figures"), mustWork = FALSE)
dir.create(FIGDIR, showWarnings = FALSE, recursive = TRUE)

## Pandoc resolves an image path against the directory of the input file, and an ABSOLUTE path
## reaches LaTeX with a `.` glued to the front. Keep the absolute path for the existence check in
## R and hand pandoc the relative one.
fig_rel <- function(f) file.path(basename(FIGDIR), basename(f))

styled <- function(x, ..., widths = NULL, bold_rows = NULL, stripe_by = NULL) {
  latex_opts <- if (is.null(stripe_by)) "striped" else NULL
  linesep <- ""
  if (!is.null(stripe_by)) {
    g <- match(stripe_by, unique(stripe_by))
    linesep <- ifelse(c(diff(g), 1) != 0, "\\addlinespace", "")
    linesep[length(linesep)] <- ""
  }
  k <- kableExtra::kbl(x, linesep = linesep, ...) |>
    kableExtra::kable_styling(latex_options = latex_opts, font_size = 9)
  if (!is.null(stripe_by)) {
    g <- match(stripe_by, unique(stripe_by))
    shaded <- which(g %% 2L == 0L)
    if (length(shaded)) {
      k <- kableExtra::row_spec(k, shaded, background = "#EFEFEF")
    }
  }
  for (i in seq_along(widths)) {
    if (nzchar(widths[i])) k <- kableExtra::column_spec(k, i, width = widths[i])
  }
  if (!is.null(bold_rows)) {
    k <- kableExtra::row_spec(k, bold_rows, bold = TRUE)
  }
  k
}

emit <- function(k) cat("\n\n", as.character(k), "\n\n", sep = "")
num <- function(x, d = 1) formatC(round(x, d), format = "f", big.mark = ",", digits = d)
pct <- function(x, d = 1) paste0(num(x, d), " %")

RUNS <- list(
  regional_30 = "2026-08-26_p30_r1429_bs143",
  regional_90 = "2026-08-26_p90_r477_bs49",
  local_30 = "2026-08-26_p30_r86_bs9",
  local_90 = "2026-08-26_p90_r29_bs3"
)
DIRS <- lapply(RUNS, run_dir)
have <- function(...) all(!is.na(unlist(DIRS[c(...)])))

ok_cells <- have("regional_30", "regional_90", "local_30", "local_90")
ok_zonal <- ok_cells &&
  all(file.exists(vapply(
    DIRS[c("regional_30", "regional_90", "local_30", "local_90")],
    function(d) file.path(d, "zonal_by_polys.csv"),
    ""
  )))

CLASSES <- c("Impeded", "Diffuse", "Intensified", "Channelized")
```

```{r class-compute, include = FALSE}
## Connectivity categories of Cameron et al. (2022), on normalized current flow F = C / Cp.
##
## Written as explicit comparisons rather than `cut()` so that the open and closed boundaries
## match the paper exactly: Impeded is F <= 0.7, Diffuse is > 0.7 and < 1.3, Intensified is
## >= 1.3 and <= 1.7, Channelized is > 1.7.
classify_F <- function(f) {
  out <- rep(NA_character_, length(f))
  out[f <= 0.7] <- "Impeded"
  out[f > 0.7 & f < 1.3] <- "Diffuse"
  out[f >= 1.3 & f <= 1.7] <- "Intensified"
  out[f > 1.7] <- "Channelized"
  factor(out, levels = CLASSES)
}

## The 30 m regional surface is 69.2 M cells, so this is the expensive step.
##
## Normalized current is a RATIO of two per-cell currents, so numerator and denominator are
## aggregated separately and then divided. Averaging the ratio itself would be wrong: the mean of
## a ratio is not the ratio of the means, and the denominator varies within a 3x3 block.
##
## Nothing is resampled. `aggregate(fact = 3)` pads the extent to fit whole blocks, and on this
## study area that padding lands exactly on the 90 m grid, so cells are compared one for one.
compare_pair <- function(k30, k90) {
  suppressMessages(library(terra))
  c30 <- terra::aggregate(
    terra::rast(file.path(DIRS[[k30]], "cum_currmap.tif")),
    fact = 3,
    fun = "mean",
    na.rm = TRUE
  )
  f30 <- terra::aggregate(
    terra::rast(file.path(DIRS[[k30]], "flow_potential.tif")),
    fact = 3,
    fun = "mean",
    na.rm = TRUE
  )
  n30 <- c30 / f30
  n90 <- terra::rast(file.path(DIRS[[k90]], "normalized_cum_currmap.tif"))

  aligned <- isTRUE(all.equal(
    as.vector(terra::ext(n30)),
    as.vector(terra::ext(n90)),
    tolerance = 1e-6
  )) &&
    identical(dim(n30)[1:2], dim(n90)[1:2])

  ## Count only what is inside the district. Omniscape needs a buffer around the study area for
  ## the moving window, so the rasters reach well beyond the NRD -- about three cells in every
  ## four of the full extent. That margin is not a result, and leaving it in flatters the
  ## comparison: it is mostly uniform low-flow land that both pixel sizes classify identically.
  ##
  ## This stops rather than carrying on unclipped if the boundary is missing. Silently falling
  ## back is how the first version of this report came to quote agreement over the full extent
  ## while describing it as the district: the numbers looked plausible and nothing flagged them.
  sa <- study_area_path()
  if (is.na(sa)) {
    stop(
      "district boundary not found; refusing to report statistics over the unclipped extent",
      call. = FALSE
    )
  }
  bnd <- terra::vect(sa)
  if (!terra::same.crs(bnd, n90)) {
    bnd <- terra::project(bnd, terra::crs(n90))
  }
  n30 <- terra::mask(terra::crop(n30, bnd), bnd)
  n90 <- terra::mask(terra::crop(n90, bnd), bnd)

  v30 <- terra::values(n30)[, 1]
  v90 <- terra::values(n90)[, 1]
  ok <- is.finite(v30) & is.finite(v90)
  a <- v30[ok]
  b <- v90[ok]

  k_a <- classify_F(a)
  k_b <- classify_F(b)
  tab <- table(r30 = k_a, r90 = k_b)
  agree <- sum(diag(tab)) / sum(tab)
  pe <- sum(rowSums(tab) / sum(tab) * colSums(tab) / sum(tab))

  list(
    aligned = aligned,
    n = sum(ok),
    tab = tab,
    agree = 100 * agree,
    kappa = (agree - pe) / (1 - pe),
    mean30 = mean(a),
    mean90 = mean(b),
    medad = median(abs(b - a)),
    r30 = n30,
    r90 = n90
  )
}

cm <- list()
if (ok_cells) {
  cm$regional <- compare_pair("regional_30", "regional_90")
  cm$local <- compare_pair("local_30", "local_90")
}
```

```{r zonal-compute, include = FALSE}
zonal_of <- function(run_key) {
  f <- file.path(DIRS[[run_key]], "zonal_by_polys.csv")
  if (is.na(DIRS[[run_key]]) || !file.exists(f)) {
    return(NULL)
  }
  z <- utils::read.csv(f, stringsAsFactors = FALSE)
  stats::setNames(z$mean_norm_cum_curr, z$zone)
}

## A handful of high-alpine units carry no current at all in either run; 0 to 0 is no basis for a
## percentage, so those zones are dropped rather than allowed to propagate a NaN into a median.
zonal_metrics <- function(a_key, b_key) {
  a <- zonal_of(a_key)
  b <- zonal_of(b_key)
  if (is.null(a) || is.null(b)) {
    return(NULL)
  }
  k <- intersect(names(a), names(b))
  a <- a[k]
  b <- b[k]
  ok <- is.finite(a) & is.finite(b) & a != 0
  a <- a[ok]
  b <- b[ok]
  d <- 100 * (b - a) / a
  list(
    n = length(a),
    median_abs = median(abs(d)),
    mean_abs = mean(abs(d)),
    over10 = 100 * mean(abs(d) > 10),
    flip = 100 * mean((a > 1) != (b > 1)),
    rho = cor(rank(a), rank(b)),
    a = a,
    b = b
  )
}

zm <- list()
if (ok_zonal) {
  zm$reg_res <- zonal_metrics("regional_30", "regional_90")
  zm$loc_res <- zonal_metrics("local_30", "local_90")
}
```

{{< pagebreak >}}

# Why resolution matters {#sec-resolution}

Quesnel was produced at both 30 m and 90 m pixel size, and a decision was taken in spring 2026 that 90 m is sufficient going forward, for further Quesnel runs and for subsequent analyses.
It was a judgement at the time, made without a formal comparison.
This comparison is formalized here.

Omniscape solves a circuit problem repeatedly in a moving window, and the window radius is given in pixels rather than in metres [@Landau:2021].
Pixel size therefore fixes two things at once: the ground distance the window spans, and how many solves a run needs to cover the study area.
Cost per solve rises faster than window area -- roughly `nodes^1.8` on this study area -- so the two compound.
This coupling is why the algorithm carries a block-size shortcut at all.
@McRae:2016 found centring the window on every pixel computationally prohibitive and solved whole blocks of pixels at once, which sped the work up without coarsening the resistance data underneath it.
Choosing a pixel size is thus a choice about what the model can resolve and about whether it finishes, together.

Connectivity surfaces of this kind are largely robust to how they are gridded: models of the same landscape built at different pixel sizes and extents correlate closely, even where the conservation plans eventually drawn from them differ [@Dertien:2023].
That is the basis for expecting 90 m results to be broadly comparable with 30 m ones.
The category cut-offs fit that expectation rather than straining it.
They were derived by @McRae:2016 for the Pacific Northwest at 180 m, and @Cameron:2022 applied the same values in California at 90 m, having tested 54 alternative sets and found the originals still described their landscape best.
The regional run compared here is 90 m with a 42.9 km search radius, against Cameron's 50 km, so it sits inside the range these cut-offs have been used at and is finer than the pixel size they were derived at.
That is a reason to expect 90 m to hold up, not evidence that it does, so the rest of this report tests it.

The reason for wanting 90 m is computational effort.
The 30 m series is 208 h of Omniscape against 12 h for the 90 m series, a factor of about seventeen, and nearly all of it is the one regional run, whose window covers nine times the area of the 90 m one.
A district's run takes about half a day at 90 m and over a week at 30 m, displacing other work on the shared machines for that period.
Memory is not the constraint at 90 m, where the regional run was measured at 33.2 GB resident and the local at 3.0 GB; data preparation rather than Omniscape sets the memory requirement.

```{r}
#| label: tbl-compute
#| tbl-cap: "Omniscape wall clock for the corrected Quesnel runs. Worker counts differ by run; these are elapsed times, not CPU hours."
#| results: asis
data.frame(
  Run = c("Regional, 30 m", "Regional, 90 m", "Local, 30 m", "Local, 90 m"),
  Radius = c("1429 px", "477 px", "86 px", "29 px"),
  Workers = c("64", "16", "24", "8"),
  Elapsed = c("196 h 44 min", "11 h 08 min", "11 h 28 min", "57 min"),
  check.names = FALSE,
  stringsAsFactors = FALSE
) |>
  styled(booktabs = TRUE, row.names = FALSE, align = "lrrr") |>
  emit()
```

{{< pagebreak >}}

# Connectivity categories {#sec-classes}

The results are put through a reclassification into connectivity categories, and categories are a harder test of pixel size than the underlying maps.
A conclusion drawn from a continuous map tolerates a small shift in value.
A conclusion drawn from categories does not, because a location sitting near a cut-off changes category outright.

@Cameron:2022 sort each location into a category by comparing two maps: the movement the model actually predicts there, and the movement the same place would carry if nothing were in the way.

The ratio of the two is what the categories are defined on.
A ratio near 1 means a place carries about as much movement as it could.
Well above 1 means movement is being funnelled through it.
Below 1 means something is getting in the way.
The ratio is written `F` in the table below, following the paper.

```{r}
#| label: tbl-thresholds
#| tbl-cap: "Category thresholds on normalized current flow, after Cameron et al. (2022)."
#| results: asis
data.frame(
  Category = CLASSES,
  `Normalized current flow` = c(
    "F <= 0.7",
    "0.7 < F < 1.3",
    "1.3 <= F <= 1.7",
    "F > 1.7"
  ),
  Meaning = c(
    "Less flow than the landscape could support",
    "About as much flow as expected",
    "More flow than expected, with few constraints",
    "Much more flow than expected; movement concentrated"
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
) |>
  styled(
    booktabs = TRUE,
    row.names = FALSE,
    align = "lll",
    widths = c("2.6cm", "3.4cm", "8.6cm")
  ) |>
  emit()
```

## The categories not computed here {#sec-missing-categories}

The framework additionally splits low-flow Diffuse and Impeded land into Impeded, Limited and Permeable, using a human modification index to separate them.
**That split is outside what is compared here.**
The pipeline does not currently produce such an index, so it is not something these outputs determine on their own.
It could be built: the roads and other anthropogenic footprint layers the pipeline already assembles are the same kind of input such an index is made from.
It was not part of the design of this stage of the work.

What was handed off was the Omniscape output rasters, so whether those three categories were derived downstream is not established here.
The four categories above are the ones the outputs fix by themselves, and they are what this comparison covers.

```{r}
#| label: tbl-confusion-regional
#| tbl-cap: "Regional run: connectivity category at 30 m against 90 m, as a percentage of all classified cells. The diagonal is agreement."
#| eval: !expr isTRUE(ok_cells)
#| results: asis
conf_tbl <- function(m) {
  p <- 100 * prop.table(m$tab)
  d <- as.data.frame.matrix(round(p, 2))
  d[] <- lapply(d, function(x) num(x, 2))
  cbind(`30 m \\ 90 m` = rownames(d), d)
}
emit(styled(
  conf_tbl(cm$regional),
  booktabs = TRUE,
  row.names = FALSE,
  align = "lrrrr"
))
```

```{r}
#| label: tbl-confusion-local
#| tbl-cap: "Local run: connectivity category at 30 m against 90 m, as a percentage of all classified cells."
#| eval: !expr isTRUE(ok_cells)
#| results: asis
emit(styled(
  conf_tbl(cm$local),
  booktabs = TRUE,
  row.names = FALSE,
  align = "lrrrr"
))
```

```{r}
#| label: tbl-perclass
#| tbl-cap: "Each category's share of the district at both resolutions, and the share of its 30 m cells that keep the category at 90 m. Extent is how much land the category covers; retention is whether it is the same land."
#| eval: !expr isTRUE(ok_cells)
#| results: asis
per_class <- function(lab, m) {
  tab <- m$tab
  do.call(
    rbind,
    lapply(CLASSES, function(lv) {
      a30 <- sum(tab[lv, ])
      a90 <- sum(tab[, lv])
      data.frame(
        Scale = lab,
        Category = lv,
        `Extent, 30 m` = pct(100 * a30 / sum(tab), 2),
        `Extent, 90 m` = pct(100 * a90 / sum(tab), 2),
        `Extent ratio` = num(a90 / a30, 3),
        `Cells retained` = pct(100 * tab[lv, lv] / a30, 1),
        check.names = FALSE,
        stringsAsFactors = FALSE
      )
    })
  )
}
tbl <- rbind(per_class("Regional", cm$regional), per_class("Local", cm$local))
emit(styled(
  tbl,
  booktabs = TRUE,
  row.names = FALSE,
  align = "llrrrr",
  stripe_by = tbl$Scale
))
```

Agreement is `r if (ok_cells) num(cm$regional$agree, 1) else "--"`% for the regional run and `r if (ok_cells) num(cm$local$agree, 1) else "--"`% for the local one.
Some of that is luck, because most of the district falls into one or two common categories, so two maps agreeing at random would still coincide fairly often.
Cohen's kappa [@Cohen:1960] discounts the luck, scoring 0 when agreement is no better than chance and 1 when it is exact.
Kappa is `r if (ok_cells) num(cm$regional$kappa, 2) else "--"` for the regional run and `r if (ok_cells) num(cm$local$kappa, 2) else "--"` for the local one.
On the kappa bands of @Landis:1977, both count as substantial agreement.

Three things follow.

**The extents are stable.**
Every category covers close to the same share of the district at either pixel size.
The largest departure is Impeded in the regional run, which grows by 11%; every other category is within 10%.
A tally of how much land falls in each category is nearly unchanged by pixel size.

**Channelized land is both stable in extent and largely the same land.**
It covers 12.6% of the district at 30 m and 13.0% at 90 m in the regional run, and `r if (ok_cells) num(100 * cm$regional$tab["Channelized", "Channelized"] / sum(cm$regional$tab["Channelized", ]), 0) else "--"`% of its locations keep the category at the coarser pixel size.
This matters more than the other categories because it is what the framework is for: Channelized areas are the constricted places where conservation or restoration is most consequential.

**Intensified is the exception.**
It retains only `r if (ok_cells) num(100 * cm$regional$tab["Intensified", "Intensified"] / sum(cm$regional$tab["Intensified", ]), 0) else "--"`% of its locations in the regional run and `r if (ok_cells) num(100 * cm$local$tab["Intensified", "Intensified"] / sum(cm$local$tab["Intensified", ]), 0) else "--"`% in the local one.
It is a narrow band -- a ratio between 1.3 and 1.7 -- holding two to three percent of the district and sitting directly between Diffuse below and Channelized above.
A small change in the ratio therefore carries a location out of it in one direction or the other.
Its *extent* is reproduced well; *which* cells make it up is not.
Conclusions that depend on Intensified land at specific locations should be drawn at 30 m, or treated as provisional at 90 m.

```{r}
#| label: fig-maps
#| fig-cap: "Quesnel NRD at both pixel sizes. Each row is the same comparison; the left column is 30 m coarsened onto the 90 m grid, the right column is 90 m as produced. The category rows are what the results are read from, and they are close to indistinguishable at map scale."
#| eval: !expr isTRUE(ok_cells)
#| out-width: "100%"
f <- file.path(FIGDIR, "resolution-maps.png")
if (!file.exists(f)) {
  suppressMessages(library(terra))
  ## `compare_pair()` already clipped these to the district, so the maps and the tables are
  ## drawn from exactly the same surfaces.

  cat_cols <- c(
    Impeded = "#B2182B",
    Diffuse = "#F7F7F7",
    Intensified = "#92C5DE",
    Channelized = "#2166AC"
  )
  as_cat <- function(r) {
    k <- terra::classify(
      r,
      rbind(c(-Inf, 0.7, 1), c(0.7, 1.3, 2), c(1.3, 1.7, 3), c(1.7, Inf, 4)),
      include.lowest = TRUE
    )
    levels(k) <- data.frame(id = 1:4, category = CLASSES)
    k
  }

  grDevices::png(f, width = 1800, height = 1880, res = 170)
  op <- graphics::par(mfrow = c(4, 2), mar = c(0.6, 1, 2.2, 7))
  rng <- c(0, 2)
  for (sc in c("regional", "local")) {
    lab <- if (sc == "regional") "Regional" else "Local"
    m <- cm[[sc]]
    a <- m$r30
    b <- m$r90
    terra::plot(
      terra::clamp(a, rng[1], rng[2], values = TRUE),
      main = paste(lab, "- movement ratio, 30 m"),
      axes = FALSE,
      range = rng,
      mar = c(0.6, 1, 2.2, 7)
    )
    terra::plot(
      terra::clamp(b, rng[1], rng[2], values = TRUE),
      main = paste(lab, "- movement ratio, 90 m"),
      axes = FALSE,
      range = rng,
      mar = c(0.6, 1, 2.2, 7)
    )
    terra::plot(
      as_cat(a),
      main = paste(lab, "- categories, 30 m"),
      axes = FALSE,
      col = cat_cols,
      mar = c(0.6, 1, 2.2, 7)
    )
    terra::plot(
      as_cat(b),
      main = paste(lab, "- categories, 90 m"),
      axes = FALSE,
      col = cat_cols,
      mar = c(0.6, 1, 2.2, 7)
    )
  }
  graphics::par(op)
  grDevices::dev.off()
}
include_graphics(fig_rel(f), rel_path = FALSE, error = FALSE)
```

{{< pagebreak >}}

# Zonal summaries {#sec-zonal}

Zonal summaries are the other product the analysis delivers: one mean normalized current per Landscape Unit by BEC zone combination, and per biodiversity feature class.

```{r}
#| label: tbl-zonal
#| tbl-cap: "Zonal summaries at 30 m against 90 m, both corrected. Percentages are the change from the 30 m value to the 90 m value, over the zones both runs report."
#| eval: !expr isTRUE(ok_zonal)
#| results: asis
zrow <- function(lab, m) {
  data.frame(
    Scale = lab,
    Zones = num(m$n, 0),
    `Median abs. change` = pct(m$median_abs, 2),
    `Mean abs. change` = pct(m$mean_abs, 2),
    `Zones moving > 10%` = pct(m$over10, 1),
    `Zones crossing 1` = pct(m$flip, 1),
    `Rank correlation` = num(m$rho, 3),
    check.names = FALSE,
    stringsAsFactors = FALSE
  )
}
emit(styled(
  rbind(zrow("Regional", zm$reg_res), zrow("Local", zm$loc_res)),
  booktabs = TRUE,
  row.names = FALSE,
  align = "lrrrrrr"
))
```

The ordering of zones is what survives best: a planner ranking Landscape Units by mean normalized current gets very nearly the same ordering from either grid.
Individual values move more, and a small number of zones fall on opposite sides of 1.

Of the regional zones that cross, the median sits `r if (ok_zonal) num(median(abs(zm$reg_res$a[(zm$reg_res$a > 1) != (zm$reg_res$b > 1)] - 1)), 3) else "--"` from the threshold, against `r if (ok_zonal) num(median(abs(zm$reg_res$a[(zm$reg_res$a > 1) == (zm$reg_res$b > 1)] - 1)), 3) else "--"` for the zones that do not.
They change category because they were already sitting on the line.

```{r}
#| label: fig-zonal
#| fig-cap: "Zonal means at 30 m against 90 m, one point per zone. The diagonal is exact agreement; grey lines mark the value of 1. Red points are zones the two grids place on opposite sides of it, and they lie close to the lines."
#| eval: !expr isTRUE(ok_zonal)
#| out-width: "100%"
f <- file.path(FIGDIR, "resolution-zonal-scatter.png")
if (!file.exists(f)) {
  grDevices::png(f, width = 2000, height = 1050, res = 190)
  op <- graphics::par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.6, 1))
  for (nm in c("Regional", "Local")) {
    m <- if (nm == "Regional") zm$reg_res else zm$loc_res
    lim <- range(c(m$a, m$b), finite = TRUE)
    lim[1] <- max(lim[1], 1e-3)
    plot(
      m$a,
      m$b,
      log = "xy",
      xlim = lim,
      ylim = lim,
      pch = 19,
      cex = 0.5,
      col = grDevices::adjustcolor(ifelse((m$a > 1) != (m$b > 1), "#C0392B", "#2C3E50"), 0.65),
      xlab = "30 m (aggregated to 90 m)",
      ylab = "90 m",
      main = paste0(nm, ": mean normalized current by zone")
    )
    graphics::abline(0, 1, col = "#2980B9", lwd = 1.6)
    graphics::abline(h = 1, v = 1, col = "grey65", lty = 2)
  }
  graphics::par(op)
  grDevices::dev.off()
}
include_graphics(fig_rel(f), rel_path = FALSE, error = FALSE)
```

# Limitations {#sec-limits}

**Measured on Quesnel only.**
Quesnel is the only district with both series, so it is the only place this comparison can be made.
A district with different terrain, size or disturbance history could be more sensitive to pixel size, and that is not measured here.

**Neither grid is truth.**
The 30 m run is finer and closer to the resolution of the source data, so where the two differ it is the better estimate.
This measures how far apart they are, not that they agree.

**Category extents transfer; individual Intensified cells do not.**
Land tallies by category, Channelized extent, and zonal rankings all survive the change of grid.
The membership of the narrow Intensified band does not, and a map read cell by cell in that category should not be treated as interchangeable between resolutions.

**Three of the framework's categories are out of scope.**
Separating low-flow land into Impeded, Limited and Permeable needs a human modification index, which this pipeline does not currently produce, though it could be derived from footprint data already in hand.
Those three categories are therefore not compared, and nothing here speaks to how pixel size would affect them.

{{< pagebreak >}}

# Appendix A: How the comparison was made {#sec-method .unnumbered}

Four corrected runs exist for Quesnel: a regional and a local run at each resolution.
The comparison is between resolutions within each scale, both runs corrected, so the grid is the only thing that differs.

```{r}
#| label: tbl-runs
#| tbl-cap: "The runs compared. Radius is stored in pixels, so the same physical distance is a different number at each resolution; the two are within one 90 m pixel of each other."
#| results: asis
data.frame(
  Scale = c("Regional", "Regional", "Local", "Local"),
  Resolution = c("30 m", "90 m", "30 m", "90 m"),
  `Radius (px)` = num(c(1429, 477, 86, 29), 0),
  `Radius (m)` = num(c(1429 * 30, 477 * 90, 86 * 30, 29 * 90), 0),
  Run = c(RUNS$regional_30, RUNS$regional_90, RUNS$local_30, RUNS$local_90),
  check.names = FALSE,
  stringsAsFactors = FALSE
) |>
  styled(
    booktabs = TRUE,
    row.names = FALSE,
    align = "llrrl",
    stripe_by = c(1, 1, 2, 2)
  ) |>
  emit()
```

The radii are derived from the corrected interpatch distances in metres and converted to pixels per resolution, so the regional runs are 42,870 m and 42,930 m and the local runs 2,580 m and 2,610 m.
They differ by less than one 90 m pixel, so this compares two grids resolving the same physical neighbourhood rather than two different analyses.

**Only the ratio can be compared between pixel sizes.**
The two underlying maps report movement per pixel, and a 90 m pixel covers nine times the ground of a 30 m one, so their values are not on the same footing.
The ratio cancels that out, and is what the categories are defined on.

To compare them, the 30 m results are coarsened onto the 90 m grid.
The two underlying maps are averaged separately and the ratio taken afterwards, rather than averaging the ratio itself.
Averaging the ratio would distort it, because the amount of movement a place could carry varies within each 90 m block.

```{r}
#| label: tbl-alignment
#| tbl-cap: "Grid alignment. Aggregating the 30 m grid by a factor of three lands exactly on the 90 m grid, so nothing is interpolated and cells are compared one for one."
#| eval: !expr isTRUE(ok_cells)
#| results: asis
data.frame(
  Scale = c("Regional", "Local"),
  `Grids identical after aggregation` = c(
    ifelse(cm$regional$aligned, "yes", "NO"),
    ifelse(cm$local$aligned, "yes", "NO")
  ),
  `Cells compared` = c(
    format(cm$regional$n, big.mark = ","),
    format(cm$local$n, big.mark = ",")
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
) |>
  styled(booktabs = TRUE, row.names = FALSE, align = "llr") |>
  emit()
```

{{< pagebreak >}}

{{< pagebreak >}}

# Appendix B: Reproducing this report {#sec-reproducing .unnumbered}

This report is not a pipeline target.
It reads the finished Omniscape outputs directly, so it renders without a live `targets` store:

- `cum_currmap.tif`, `flow_potential.tif` and `normalized_cum_currmap.tif` from each of the four
corrected runs;
- `zonal_by_polys.csv` from those same four runs.

It resolves both output layouts, so it renders against a clone from before Quesnel's artefacts moved under `quesnel/` as well as after.

Render with:

```bash
QUARTO_R=/path/to/R quarto render reports/resolution-equivalence.qmd \
  --to pdf --output-dir ../Outputs
```

```{r}
#| label: tbl-session
#| tbl-cap: "Environment."
#| results: asis
si <- utils::sessionInfo()
data.frame(
  Component = c("R", "Platform", "OS", "terra", "GDAL", "GEOS", "PROJ"),
  Version = c(
    paste(si$R.version$major, si$R.version$minor, sep = "."),
    si$platform,
    if (is.null(si$running)) "unknown" else si$running,
    as.character(utils::packageVersion("terra")),
    unname(sf::sf_extSoftVersion()["GDAL"]),
    unname(sf::sf_extSoftVersion()["GEOS"]),
    unname(sf::sf_extSoftVersion()["PROJ"])
  ),
  stringsAsFactors = FALSE
) |>
  styled(booktabs = TRUE, row.names = FALSE) |>
  emit()
```

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::
